---
title: "Guidance efficiency model for LGR smolt abundance"
subtitle: "MY2025 steelhead — methods and results"
author: "David Smith, IDFG Wild Salmon and Steelhead Program"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: false
    fig_width: 10
    fig_height: 5
    latex_engine: xelatex
geometry: margin=1in
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(lubridate)
library(scales)
library(knitr)
library(ISOweek)
library(smoltEASE)
library(coda)

input_dir <- "C:/Users/david.smith/OneDrive - State of Idaho/Desktop/SCRAPI/SCRAPI2/Tim Meet"

```

SCRAPI estimates wild smolt passage at Lower Granite Dam (LGR) by expanding
the daily bypass trap sample by guidance efficiency (GE), the fraction of
outmigrants that enter the juvenile bypass system:

$$N_i = \frac{c_i}{r_i \cdot \widehat{GE}_i}$$

where $c_i$ is the daily trap count, $r_i$ the sampling rate, and $\widehat{GE}_i$
the guidance efficiency on day $i$. GE is the most consequential input. It is strongly negatively correlated with
spill, and under the 2020 Record of Decision LGR now runs at 80 to 85% spill
through the April to May migration peak. The GE values SCRAPI currently uses
come from Sandford and Smith (2002), developed under pre-ROD low-spill
conditions. At 80 to 85% spill that framework extrapolates well outside the
range it was fit on.

---

## Why current GE is likely too high

```{r load-data}
STHD.FPC <- read.csv(file.path(input_dir, "MY2025STHDsmolts_FPCdata_strata_added.csv"))
STHD.FPC$SampleEndDate <- format(as.Date(STHD.FPC$SampleEndDate, "%m/%d/%Y"), "%m/%d/%Y")

pit <- read_csv(file.path(input_dir, "MY25_STHD.csv")) |>
  mutate(site     = str_extract(`Site Name`, "^[A-Z]+"),
         obs_date = as_date(parse_date_time(`Obs Time Value`,
                            orders = c("mdy HM", "mdy HMS", "mdy HMSp")))) |>
  filter(site %in% c("GRJ", "GRS"),
         obs_date >= ymd("2025-03-01"), obs_date < ymd("2025-08-01")) |>
  group_by(`Tag Code`, site) |> slice_min(obs_date, with_ties = FALSE) |> ungroup() |>
  mutate(Week = isoweek(obs_date)) |> count(Week, site) |>
  pivot_wider(names_from = site, values_from = n, values_fill = 0) |>
  rename(n_grj = GRJ, n_grs = GRS) |>
  mutate(GE_pit_d50 = replace_na(n_grj / (n_grj + n_grs / 0.50), 0),
         GE_pit_d25 = replace_na(n_grj / (n_grj + n_grs / 0.25), 0))

dart <- read_csv(file.path(input_dir, "DART Spill.csv")) |>
  mutate(date         = mdy(date),
         spill_per    = as.numeric(spill_kcfs) / as.numeric(outflow_kcfs) * 100,
         outflow_kcfs = as.numeric(outflow_kcfs)) |>
  filter(date >= ymd("2025-03-01"), date < ymd("2025-08-01")) |>
  mutate(Week = isoweek(date)) |>
  group_by(Week) |>
  summarise(spill_per    = replace_na(mean(spill_per, na.rm = TRUE), 0),
            outflow_kcfs = mean(outflow_kcfs, na.rm = TRUE), .groups = "drop")
```

```{r ge-plot, fig.cap="Weekly GE for MY2025 wild steelhead. NOAA GE holds near 13% through peak spill while spill-based alternatives fall to single digits."}
wk_date <- function(w) ISOweek2date(paste0("2025-W", str_pad(w, 2, pad = "0"), "-1"))

css_ge  <- dart |>
  mutate(GE_css = plogis(0.85 + 0.01066*(outflow_kcfs - 24) - 0.08227*spill_per - 0.34165)) |>
  select(Week, GE_css)
ge_noaa <- STHD.FPC |> group_by(Week) |>
  summarise(GE_noaa = mean(GuidanceEfficiency, na.rm = TRUE), .groups = "drop")

ge_lines <- ge_noaa |>
  left_join(css_ge, by = "Week") |>
  left_join(pit |> select(Week, GE_pit_d50, GE_pit_d25), by = "Week") |>
  mutate(week_start = wk_date(Week)) |>
  pivot_longer(c(GE_noaa, GE_css, GE_pit_d50, GE_pit_d25),
               names_to = "method", values_to = "GE") |>
  mutate(method = recode(method,
    GE_noaa    = "Current (NOAA) GE",
    GE_css     = "McCann et al. (2023) logit",
    GE_pit_d50 = "PIT GE (det = 0.50)",
    GE_pit_d25 = "PIT GE (det = 0.25)"))

xr   <- range(ge_lines$week_start[!is.na(ge_lines$GE)])
bars <- dart |> mutate(week_start = wk_date(Week)) |>
  filter(week_start >= xr[1], week_start <= xr[2])

pal <- c("Current (NOAA) GE"          = "black",
         "McCann et al. (2023) logit"  = "#1a9641",
         "PIT GE (det = 0.50)"         = "#2166ac",
         "PIT GE (det = 0.25)"         = "#d01c8b")

ggplot() +
  geom_col(data = bars, aes(week_start, spill_per/100), fill = "steelblue",
           alpha = 0.18, width = 5) +
  geom_line(data  = ge_lines, aes(week_start, GE, color = method),
            linewidth = 0.9, na.rm = TRUE) +
  geom_point(data = ge_lines, aes(week_start, GE, color = method),
             size = 1.8, na.rm = TRUE) +
  scale_color_manual(values = pal) +
  scale_y_continuous("Guidance efficiency", labels = percent_format(accuracy = 1),
                     limits = c(0, 1),
                     sec.axis = sec_axis(~., name = "Spill proportion",
                                         labels = percent_format(accuracy = 1))) +
  coord_cartesian(xlim = xr) +
  labs(x = NULL, color = NULL,
       title = "Weekly GE comparison, MY2025 wild steelhead",
       subtitle = "Bars = weekly mean spill. Multistate model GE not shown here.") +
  theme_bw(base_size = 14) + theme(legend.position = "bottom")

```

All four lines in this figure are estimates of what GE might look like under
current conditions, not measurements of it. The NOAA line uses the Sandford
and Smith (2002) fixed estimates. The McCann line applies their spill-based
logit regression. The two PIT lines compute GE directly from the GRJ (Lower Granite Juvenile Bypass) and GRS (Lower Granite spillbay1 array)
first-detection counts, but to do that you need to know the GRS detection
probability — the chance a spillway fish actually gets picked up by the bay-1
antenna. The two curves bracket it at assumed values of 50% and 25%, which is
why they're labeled "det = 0.50" and "det = 0.25." None of these approaches
actually estimates that detection probability; they all assume it or ignore it.

That is the core problem. **GRS monitors only 1 of 8 spillbays, and under
post-ROD high spill the fraction of fish using bay 1 and the antenna's
intrinsic efficiency are both unknown and probably time-varying.** Without
knowing the GRS detection probability, any direct PIT-based GE calculation
is just a bracket, not an estimate. The multistate model below estimates that
detection probability jointly with GE from the PIT data themselves, using
downstream recoveries at Little Goose juvenile bypass facility (GOJ) as the identifying information.

---

## The multistate model

A stratum-level multistate mark-recapture model on first-detection histories
of PIT-tagged fish at three arrays. Per weekly stratum $s$ it jointly estimates:

| Parameter | Quantity |
|---|---|
| $\psi_s$ | GE — P(fish takes the bypass route) |
| $p_s$ | GRS detection — P(spillway fish detected at bay 1) |
| $\phi_s$ | P(fish survives LGR to LGS and is detected at the GOJ bypass) |

Each detected fish falls into one of five histories. Conditioned on detection:

| History | Route | Probability |
|---|---|---|
| GRJ, not GOJ | Bypass, missed downstream | $\psi(1-\phi)$ |
| GRJ and GOJ | Bypass, seen downstream | $\psi\phi$ |
| GRS, not GOJ | Spillway, detected at bay 1, missed downstream | $(1-\psi)\,p\,(1-\phi)$ |
| GRS and GOJ | Spillway, detected at bay 1, seen downstream | $(1-\psi)\,p\,\phi$ |
| No LGR, GOJ | Spillway, missed at bay 1, seen downstream | $(1-\psi)(1-p)\phi$ |

The fifth history — missed at both LGR arrays, recovered at Little Goose —
is what identifies $p$ and corrects GE for fish GRS never saw.

Each parameter is a hierarchical logistic regression on spill. $\psi_s$ is
nested in a parent collapse-stratum mean, so sparse weeks borrow strength
rather than producing high-leverage near-zero estimates:

$$\text{logit}(\psi_s) \sim N\!\big(\mu_{\text{strat}[g(s)]} + \beta \cdot \text{spill}^{\text{LGR}}_s,\ \sigma^2_\psi\big)$$
$$\text{logit}(p_s) \sim N\!\big(\alpha_p + \beta_p \cdot \text{spill}^{\text{LGR}}_s,\ \sigma^2_p\big)$$
$$\text{logit}(\phi_s) \sim N\!\big(\alpha_\phi + \beta_\phi \cdot \text{spill}^{\text{LGS}}_s,\ \sigma^2_\phi\big)$$
$$\mu_{\text{strat}[g]} \sim N(\alpha,\ \sigma^2_{\text{strat}})$$

where $g(s)$ maps weekly stratum $s$ to its parent collapse stratum
(adjacent trapping weeks combined to reach roughly 100 genotyped fish per stratum), $\alpha$ is
the global mean logit(GE), $\beta$ is the spill slope (constrained $\le 0$,
since spill can only lower GE), and $\mu_{\text{strat}[g]}$ is the
collapse-stratum mean toward which sparse weeks are pooled. $\phi$ is modeled
on LGS spill since Little Goose is the downstream detection site. $\sigma^2_\psi$
and $\sigma^2_{\text{strat}}$ control within-stratum and between-stratum
variability on the logit scale respectively.

The observed history counts follow a multinomial likelihood conditioned on
the total fish seen per stratum, evaluated only for strata with data:

$$n[s, 1{:}5] \mid N_{\text{seen}}[s] \sim \text{Multinomial}\big(N_{\text{seen}}[s],\ \pi_{\text{obs}}[s, 1{:}5]\big)$$

where $n[s, 1{:}5]$ is the vector of observed fish counts in each of the five
detection histories for stratum $s$ (i.e. how many fish were seen at GRJ only,
GRJ and GOJ, GRS only, GRS and GOJ, and GOJ only), and $N_{\text{seen}}[s] =
\sum_{k=1}^{5} n[s,k]$ is the total number of fish detected anywhere in that stratum, where $k$ indexes the five detection histories.

$\pi_{\text{obs}}[s, k]$ are the normalized cell probabilities. The raw cell
probabilities $\pi_{\text{raw}}[s, k]$ from the history table above do not sum
to 1 because there is a sixth outcome — a fish that passes LGR undetected and
is never seen at GOJ either. We cannot observe those fish, so the likelihood
can only be defined over the five histories we do see. The normalization
divides each raw probability by the total probability of being seen at all:

$$p_{\text{seen}}[s] = \sum_{k=1}^{5} \pi_{\text{raw}}[s,k]$$
$$\pi_{\text{obs}}[s, k] = \frac{\pi_{\text{raw}}[s, k]}{p_{\text{seen}}[s]}$$

This is what the JAGS line `pi_obs[s, k] <- pi_raw[s, k] / p_seen[s]` does. The unobservable sixth outcome cancels
out of the likelihood entirely.

The $\phi$ prior is derived from McCann et al. (2023) Table 8.3 LGS steelhead
coefficients, reprojected onto each migration year's standardized LGS spill
scale, so the same McCann raw coefficients translate correctly to whatever
spill conditions that year saw.

### Full annotated JAGS model

```
model {
  eps       <- 1.0E-9
  tau_psi   <- pow(sigma_psi,   -2)
  tau_strat <- pow(sigma_strat, -2)
  tau_p     <- pow(sigma_p,     -2)
  tau_phi   <- pow(sigma_phi,   -2)

  for (s in 1:S) {
    logit_psi[s] ~ dnorm(mu_strat[parent[s]] + beta * lgr_spill_std[s], tau_psi)
    psi[s] <- ilogit(logit_psi[s])        # GE for stratum s

    logit_p[s] ~ dnorm(alpha_p + beta_p * lgr_spill_std[s], tau_p)
    p[s] <- ilogit(logit_p[s])            # P(spillway fish seen at GRS bay 1)

    logit_phi[s] ~ dnorm(alpha_phi + beta_phi * lgs_spill_std[s], tau_phi)
    phi[s] <- ilogit(logit_phi[s])        # P(surviving LGR to LGS and detected at GOJ bypass)

    pi_raw[s, 1] <- psi[s] * (1 - phi[s])               + eps  # GRJ, not GOJ
    pi_raw[s, 2] <- psi[s] * phi[s]                     + eps  # GRJ and GOJ
    pi_raw[s, 3] <- (1 - psi[s]) * p[s] * (1 - phi[s])  + eps  # GRS, not GOJ
    pi_raw[s, 4] <- (1 - psi[s]) * p[s] * phi[s]        + eps  # GRS and GOJ
    pi_raw[s, 5] <- (1 - psi[s]) * (1 - p[s]) * phi[s]  + eps  # no LGR, GOJ (identifies p)

    p_seen[s] <- pi_raw[s,1] + pi_raw[s,2] + pi_raw[s,3] + pi_raw[s,4] + pi_raw[s,5]
    for (k in 1:5) { pi_obs[s, k] <- pi_raw[s, k] / p_seen[s] }
  }

  for (j in 1:S_lik) {
    n[lik_idx[j], 1:5] ~ dmulti(pi_obs[lik_idx[j], 1:5], N_seen[lik_idx[j]])
  }

  for (g in 1:n_strat) { mu_strat[g] ~ dnorm(alpha, tau_strat) }

  # Weakly-informative Student-t priors following Hance et al. (2024):
  # intercepts ~ t(df=7, scale=2); slopes ~ t(df=7, scale=1)
  alpha       ~ dt(0, pow(2, -2), 7)
  beta        ~ dt(0, pow(1, -2), 7) T(, 0)   # spill can only lower GE
  sigma_psi   ~ dunif(0.05, 3)
  sigma_strat ~ dunif(0.05, 3)

  alpha_p  ~ dt(0, pow(2, -2), 7)
  beta_p   ~ dt(0, pow(1, -2), 7)
  sigma_p  ~ dunif(0.05, 3)

  # phi prior: McCann et al. (2023) Table 8.3 LGS steelhead, reprojected
  #   alpha_raw = logit(0.81) + (-0.42932)*1 = 1.0207
  #   beta_raw  = -0.07718  (per % spill)
  #   alpha_phi_mean = alpha_raw + beta_raw * mean(lgs_spill_val)
  #   beta_phi_mean  = beta_raw  * sd(lgs_spill_val)
  alpha_phi ~ dnorm(alpha_phi_mean, pow(alpha_phi_sd, -2))
  beta_phi  ~ dnorm(beta_phi_mean,  pow(beta_phi_sd,  -2)) T(, 0)
  sigma_phi ~ dunif(0.05, 1)
}
```

---

## Data and results

```{r load-pipeline-inputs}
pit_raw <- read_csv(file.path(input_dir, "MY25_STHD.csv")) |>
  mutate(site     = str_extract(`Site Name`, "^[A-Z]+"),
         obs_date = as_date(parse_date_time(`Obs Time Value`,
                            orders = c("mdy HM", "mdy HMS", "mdy HMSp")))) |>
  filter(`Rear Type Code` %in% c("W", "H", "U"),
         site %in% c("GRJ", "GRS", "GOJ"),
         obs_date >= ymd("2025-03-01"), obs_date < ymd("2025-08-01")) |>
  group_by(`Tag Code`, site) |> slice_min(obs_date, with_ties = FALSE) |> ungroup() |>
  rename(tag = `Tag Code`, det_date = obs_date)

lgr_spill <- read_csv(file.path(input_dir, "DART Spill.csv")) |>
  mutate(Date      = mdy(date),
         spill.per = replace_na(as.numeric(spill_kcfs) / as.numeric(outflow_kcfs) * 100, 0)) |>
  filter(Date >= ymd("2025-03-01"), Date < ymd("2025-08-01")) |>
  arrange(Date) |>
  select(Date, spill.per)

lgs_spill <- bind_rows(
  read_csv(file.path(input_dir, "spill.all.csv")),
  read_csv(file.path(input_dir, "spill.all2.csv")),
  read_csv(file.path(input_dir, "spill.all3.csv"))
) |>
  mutate(Date      = as_date(mdy_hm(DateTime)),
         spill.per = HourlySpill / HourlyFlow * 100) |>
  filter(Site == "LGS", Date >= ymd("2025-03-01"), Date < ymd("2025-08-01")) |>
  group_by(Date) |>
  summarise(spill.per = mean(spill.per, na.rm = TRUE), .groups = "drop")

strata_raw  <- readxl::read_xlsx(file.path(input_dir, "Smolt Strata.xlsx"), sheet = 1)
sthd_start  <- which(strata_raw[[1]] == "WeekNumber" & strata_raw[[2]] == "STHD_n")
sthd_strata <- strata_raw[(sthd_start + 1):nrow(strata_raw), c(1, 3)] |>
  set_names(c("Week", "Collapse")) |>
  filter(!is.na(Week)) |>
  mutate(Week = as.integer(Week), Collapse = as.integer(Collapse))
```

```{r show-inputs, echo=TRUE}
# PIT detection data: one row per tag per site, first detection only.
# Tags detected at GRJ (bypass), GRS (spillbay 1), or GOJ (Little Goose).
# Rear type filtered to W/H/U. det_date is the first detection date at that site.
dplyr::glimpse(pit_raw)

# LGR daily spill: Date and spill as % of total flow.
# Drives the psi (GE) and p (GRS detection) spill regressions.
dplyr::glimpse(lgr_spill)

# LGS daily spill: Date and spill as % of total flow.
# Drives the phi (GOJ survival/bypass) spill regression.
dplyr::glimpse(lgs_spill)

# Week-to-collapse-stratum map: assigns each ISO week to a coarser collapse
# stratum. Sparse weeks within the same collapse stratum borrow strength
# from each other through the nested hierarchy in the GE model.
dplyr::glimpse(sthd_strata)

```

```{r detection-summary, echo=FALSE}
# Total unique tags detected at each site across the season
site_summary <- pit_raw |>
  group_by(site) |>
  summarise(`Unique tags` = n_distinct(tag), .groups = "drop") |>
  arrange(factor(site, levels = c("GRJ", "GRS", "GOJ")))

knitr::kable(site_summary,
             caption = "Total unique PIT tags detected at each array, MY2025 wild steelhead (W/H/U).",
             col.names = c("Array", "Unique tags"),
             format.args = list(big.mark = ","))
```

```{r detection-by-week, echo=FALSE}
# Detections by week and site — shows data volume driving the model
week_summary <- pit_raw |>
  mutate(Week = lubridate::isoweek(det_date)) |>
  group_by(Week, site) |>
  summarise(n = n_distinct(tag), .groups = "drop") |>
  tidyr::pivot_wider(names_from = site, values_from = n, values_fill = 0) |>
  select(Week, any_of(c("GRJ", "GRS", "GOJ"))) |>
  arrange(Week)

knitr::kable(week_summary,
             caption = "Unique PIT tags detected per week at each array. These counts directly drive the five detection-history cells entering the multinomial likelihood.",
             format.args = list(big.mark = ","))

```

```{r pipeline, echo=TRUE}
sthd_weekly <- transform(sthd_strata, Collapse = Week)

# prep_ge_data: takes raw PIT detections at GRJ, GRS, and GOJ and assembles
# them into weekly stratum-level detection-history counts (n_h1 through n_h5)
# plus spill covariates. This is the data structure the JAGS model consumes.
# The parent_strata argument nests weeks inside collapse strata so sparse weeks
# can borrow strength from their parent group rather than being estimated alone.
ge_data  <- smoltEASE::prep_ge_data(pit_raw, sthd_weekly, lgr_spill,
                         lgs_spill_data = lgs_spill, species = "sthd",
                         goj_site = "GOJ", parent_strata = sthd_strata)

# ge_data: one row per weekly stratum. n_h1-n_h5 are the five detection-history
# counts that go into the multinomial likelihood. spill_val and lgs_spill_val
# are the LGR and LGS mean spill percents for the stratum. parent maps each
# week to its collapse stratum for the nested hierarchy.
dplyr::glimpse(ge_data)
```

```{r history-summary, echo=FALSE}
history_summary <- ge_data |>
  summarise(
    `n_h1 (GRJ, not GOJ)` = sum(n_h1),
    `n_h2 (GRJ and GOJ)`  = sum(n_h2),
    `n_h3 (GRS, not GOJ)` = sum(n_h3),
    `n_h4 (GRS and GOJ)`  = sum(n_h4),
    `n_h5 (no LGR, GOJ)`  = sum(n_h5)
  ) |>
  tidyr::pivot_longer(everything(), names_to = "History", values_to = "Count") |>
  mutate(`% of total` = round(100 * Count / sum(Count), 1))

knitr::kable(history_summary,
             caption = "Season totals for each of the five detection histories. n_h5 (the identifying cell for p) makes up only 3.5% of detected fish, comparable to n_h4 — that scarcity is the core identifiability challenge for the GRS detection probability.",
             format.args = list(big.mark = ","))

```

```{r pipeline-cont, echo=TRUE}
# fit_ge_model: runs the Bayesian multistate JAGS model described above.
# Jointly estimates psi (GE), p (GRS bay-1 detection probability), and phi
# (probability of surviving LGR to LGS and taking the GOJ bypass) for each
# weekly stratum using the multinomial likelihood on the five detection
# histories. Returns an mcmc.list of posterior draws plus the spill
# standardization moments needed to reproject GE to daily resolution.
ge_fit   <- smoltEASE::fit_ge_model(ge_data, seed = 42)

# generate_ge_draws: converts the weekly psi posteriors into a daily GE draw
# matrix (one row per day in passageData, one column per bootstrap draw).
# For each posterior draw it recovers the stratum's process deviation
# (the part of logit(GE) not explained by spill), then substitutes that
# day's actual spill to get a daily GE value. The result is what SCRAPI2
# consumes: each bootstrap iteration draws one column, so GE uncertainty
# propagates directly into the abundance confidence intervals.
ge_draws <- smoltEASE::generate_ge_draws(ge_fit, sthd_weekly,
               pass_dates  = STHD.FPC$SampleEndDate,
               B           = 2000,
               daily_spill = lgr_spill)

# ge_draws: one row per day in passageData, one column per bootstrap draw
# (plus a SampleEndDate column). dim() confirms the shape of the matrix.
dim(ge_draws)


```

All R-hat values $\le$ `r round(max(gelman.diag(ge_fit$samples, autoburnin = FALSE, multivariate = FALSE)$psrf[,1], na.rm = TRUE), 2)` — chains converged.

```{r posterior-table}
samp_mat <- do.call(rbind, lapply(ge_fit$samples, as.matrix))

summarize_param <- function(prefix, strata_idx) {
  cols <- grep(paste0("^", prefix, "\\["), colnames(samp_mat))
  m    <- samp_mat[, cols, drop = FALSE]
  data.frame(stratum_idx = strata_idx,
             mean = colMeans(m),
             lci  = apply(m, 2, quantile, 0.025),
             uci  = apply(m, 2, quantile, 0.975))
}

strata_idx  <- ge_fit$obs_strata$stratum_idx
psi_summary <- summarize_param("psi", strata_idx); psi_summary$param <- "psi (GE)"
p_summary   <- summarize_param("p",   strata_idx); p_summary$param   <- "p (GRS detection)"
phi_summary <- summarize_param("phi", strata_idx); phi_summary$param <- "phi (survival x GOJ bypass)"

post_summary <- bind_rows(psi_summary, p_summary, phi_summary) |>
  left_join(ge_fit$obs_strata |> select(stratum_idx, stratum, spill_val, lgs_spill_val),
            by = "stratum_idx")

post_summary |>
  select(param, stratum, spill_val, mean, lci, uci) |>
  arrange(param, stratum) |>
  knitr::kable(caption = "Posterior mean and 95% CI by week",
               col.names = c("Parameter", "Week", "Spill (%)", "Mean", "2.5%", "97.5%"),
               digits = 3)

```

The final two to three strata have very few fish in the identifying history
cell (n_h5 in the single digits). Their per-week estimates are less stable
as a result, but this has negligible effect on the season total since those
weeks also contribute the least trap catch.

---

## From weekly to daily GE

The JAGS model estimates one GE value per weekly stratum — a single posterior
for each week. But SCRAPI expands trap counts daily, and GE actually moves
with daily spill. So we need to take the weekly estimate and resolve it to a
daily value. Here is how that works, step by step.

**Step 1. The spill regression.**
The model fits a line relating logit(GE) to standardized spill
($\text{std} = (\text{spill} - \bar{s}) / s_{\text{sd}}$, where $\bar{s}$ and
$s_{\text{sd}}$ are the season mean and SD of spill):

$$\text{logit}(\psi_s) = \alpha + \beta \cdot \text{std}_s + \varepsilon_s$$

$\alpha$ is the intercept (global mean logit GE), $\beta$ is the spill slope
(how much logit GE changes per unit of standardized spill), and $\varepsilon_s$
is the stratum's leftover deviation: the part of that week's GE that is not
explained by spill alone. Think of $\varepsilon_s$ as the week's fingerprint:
it captures whatever made that week's GE higher or lower than the spill
regression would predict.

**Step 2. Recover the stratum's deviation.**
Rearranging Step 1, we can back-calculate $\varepsilon_s$ from the posterior
draw of $\psi_s$:

$$\varepsilon_s = \operatorname{logit}(\psi_s) - \big(\alpha + \beta \cdot \text{std}_s\big)$$

This is the deviation we hold fixed when we move from weekly to daily.

**Step 3. Plug in that day's spill.**
For any day $d$ within stratum $s$, we substitute the day's own standardized
spill $\text{std}_d$ into the regression while keeping $\varepsilon_s$ fixed:

$$\operatorname{logit}(\widehat{GE}_d) = \alpha + \beta \cdot \text{std}_d + \varepsilon_s$$

**Step 4. The algebra simplifies.**
Substituting the expression for $\varepsilon_s$ from Step 2:

$$= \alpha + \beta \cdot \text{std}_d + \operatorname{logit}(\psi_s) - \alpha - \beta \cdot \text{std}_s$$

The $\alpha$ terms cancel, giving:

$$\boxed{\widehat{GE}_d = \operatorname{logit}^{-1}\!\big(\operatorname{logit}(\psi_s) + \beta\,(\text{std}_d - \text{std}_s)\big)}$$

So $\varepsilon_s$ is still there, it just got absorbed into $\operatorname{logit}(\psi_s)$.
The boxed equation says: start from the week's GE estimate, then adjust up or
down based on how that day's spill differs from the weekly mean, scaled by the
spill slope $\beta$. On a day where spill equals the stratum mean,
$\text{std}_d = \text{std}_s$, the adjustment is zero, and
$\widehat{GE}_d = \psi_s$ exactly.

This is done for every posterior draw independently, so the daily GE output
is a matrix of draws (one per column) that carries full uncertainty through
to the SCRAPI abundance estimate.

```{r daily-ge-plot, fig.cap="Daily GE posterior mean and 95% CI across the migration season. NOAA fixed GE shown for comparison. Bars show daily LGR spill proportion."}
ge_mat <- as.matrix(ge_draws[, -1])
ge_day <- tibble(
  date   = as.Date(ge_draws$SampleEndDate, "%m/%d/%Y"),
  mean   = rowMeans(ge_mat),
  lo     = apply(ge_mat, 1, quantile, 0.025),
  hi     = apply(ge_mat, 1, quantile, 0.975),
  NOAA   = STHD.FPC$GuidanceEfficiency
)

spill_bars <- lgr_spill |>
  rename(date = Date, spill = spill.per) |>
  filter(date >= min(ge_day$date), date <= max(ge_day$date))

ggplot() +
  geom_col(data = spill_bars, aes(date, spill / 100),
           fill = "steelblue", alpha = 0.18, width = 1) +
  geom_ribbon(data = ge_day, aes(date, ymin = lo, ymax = hi),
              fill = "#d73027", alpha = 0.15) +
  geom_line(data = ge_day, aes(date, mean, colour = "Model GE (daily)"),
            linewidth = 0.9) +
  geom_line(data = ge_day, aes(date, NOAA, colour = "NOAA fixed GE"),
            linetype = "dashed") +
  scale_colour_manual(values = c("Model GE (daily)" = "#d73027",
                                 "NOAA fixed GE"    = "grey30")) +
  scale_y_continuous("Guidance efficiency", labels = percent_format(accuracy = 1),
                     sec.axis = sec_axis(~., name = "Spill proportion",
                                         labels = percent_format(accuracy = 1))) +
  labs(x = NULL, colour = NULL,
       title = "Daily GE — model vs NOAA fixed, MY2025 wild steelhead") +
  theme_bw(base_size = 14) + theme(legend.position = "top")
```

The MCMC sampler produces thousands of plausible parameter combinations,
each a slightly different version of $(\alpha, \beta, \psi_s, \varepsilon_s)$
that is consistent with the PIT detection data. Each combination implies a
different daily GE curve for the entire season. The draw matrix stores those
curves: each column is one coherent draw from the joint posterior, i.e. one
complete daily GE time series from one plausible state of the world.

SCRAPI2's bootstrap works by randomly pulling one column per iteration. In
that iteration, instead of dividing every day's trap count by the same fixed
NOAA GE, it divides by the GE from that draw. Repeat this across 2,000
iterations and you get 2,000 abundance estimates, each using a slightly
different but internally consistent GE curve. The spread of those 2,000
estimates becomes the confidence interval.

The key consequence: weeks where GE is uncertain, primarily the peak
high-spill weeks where the PIT data are thin and the posterior is wide, get
a wider range of GE values drawn across iterations, which directly widens the
abundance CI for those weeks. Weeks where GE is well-constrained contribute
less additional uncertainty. The CI on the final season total reflects the
accumulated uncertainty across all weeks, weighted by how much trap catch each
week contributes to the expansion.

---

## Open questions

1. **GOJ identifiability.** `p` is identified only by the fifth history cell.
   Under 80 to 85% spill, GOJ detections of spillway-route fish are very thin.
   Does the McCann prior on `phi` adequately stabilize this, or does it mask
   weak identification?

2. **`phi` conflates survival and detection.** The parameter `phi` absorbs
   three distinct quantities: survival from LGR to LGS, the probability a
   fish takes the GOJ bypass route at Little Goose rather than spilling, and
   the GOJ antenna detection efficiency. These cannot be disentangled from the
   available data without additional downstream detection sites (LMJ, MCJ) or
   a richer submodel along the lines of Hance et al. (2024), who model the
   full travel-time distribution and bypass routing at Little Goose explicitly.
   Under high spill, all three components are simultaneously being squeezed:
   survival may drop slightly, the GOJ bypass rate falls because Little Goose
   is also spilling heavily, and antenna efficiency may be reduced. The McCann
   prior anchors `phi` under conditions that may not fully represent 80 to 85%
   spill, so the prior could be both stabilizing a poorly-identified parameter
   and biasing it. This is the most consequential unresolved assumption in the
   model.

3. **Bay-1 spill as a covariate for `p`.** Hance et al. (2024) handle bay-1
   operation as a hard binary multiplier: `p[t]` is forced to zero on days
   when spillbay 1 is not operating, reflecting the fact that GRS cannot
   detect fish if bay 1 is not spilling. Under pre-ROD conditions this was a
   meaningful constraint, since bay 1 operated intermittently. Under post-ROD
   operations at LGR, bay-1 spill was non-zero every day of the MY2025
   migration season (range 7 to 100% of total flow), so Hance's binary
   indicator would never trigger and has no practical effect on this analysis.

   We evaluated adding bay-1 spill percent as a continuous covariate in the
   `p` regression alongside total LGR spill. However, in MY2025 the two are
   strongly negatively correlated ($\rho = -0.98$): as post-ROD total spill
   increases, flow concentrates across more bays and the bay-1 fraction
   decreases. The two covariates carry nearly identical information about GRS
   detection probability, making their effects unidentifiable when modeled
   together. The single-covariate `p` regression on total LGR spill is
   retained. This is a feature of post-ROD operations specifically, not a
   general limitation: years or systems with more independent variation in
   bay-1 use relative to total spill could support the two-covariate version.

4. **Separate vs joint estimation.** GE is fit here as a standalone model
   and draws are plugged into SCRAPI's $1/GE$ expansion. A fully joint
   abundance-GE model would avoid the two-step decoupling and provide a
   coherent uncertainty quantification across both stages.

5. **`psi` does not track spill monotonically in the tail weeks.** Week 25
   (63.2% spill) has a posterior mean GE of 0.005, lower than week 24 (79.8%
   spill, GE = 0.010), even though spill dropped and the population-level
   slope is constrained to `beta <= 0`. This is almost certainly noise from
   the identifying cell (`n_h5` in the single digits for both weeks) rather
   than a real spill effect, and it barely moves the total: weeks 24 through
   26 contribute roughly 700 fish combined out of 164,795 in the fixed-GE
   run. Worth confirming that explanation directly, and flagging `n_h5`
   alongside any GE table used internally so a near-zero estimate built on a
   handful of tags is not read as a real signal.

6. **`phi`'s conflation of survival, routing, and detection is testable with
   existing PTAGIS data.** Downstream detections at LMJ or MCJ, if available
   for MY2025 wild steelhead, would let us bound survival from LGR to LGS
   separately from the GOJ bypass-and-detection process that `phi` currently
   absorbs whole. Even a rough side analysis there, before committing to a
   full Hance-style travel-time submodel, would show whether the McCann
   prior is stabilizing `phi` under 80 to 85% spill or biasing it.

---

## What this means for abundance

The plots above show what the model estimates GE to be. The question that
actually matters is what that does to the wild steelhead abundance estimate.
SCRAPI2 answers this by running the same trap expansion twice: once using the
fixed NOAA GE values current SCRAPI relies on, and once drawing from the
model's daily GE posterior, one draw per bootstrap iteration.

```{r load-trap, echo=TRUE}
# Trap data: one row per smolt sampled at the JBS trap. CollectionDate is
# reformatted to match the FPC date format used elsewhere in this document.
# Week and Collapse are added here since the trap file does not carry them.
STHD.trap <- read.csv(file.path(input_dir, "MY2025STHD.trapData_formatted.csv"))
STHD.trap$CollectionDate <- format(as.Date(STHD.trap$CollectionDate, "%Y-%m-%d"), "%m/%d/%Y")
STHD.trap$Week <- lubridate::isoweek(as.Date(STHD.trap$CollectionDate, "%m/%d/%Y"))
STHD.trap <- dplyr::left_join(STHD.trap, sthd_strata, by = "Week")

```

```{r scobi-fixed, echo=TRUE}
# The current fixed-GE number comes from the production tool, SCOBI::SCRAPI,
# not a SCRAPI2 reimplementation. SCOBI writes its output to CSV rather than
# returning it, so the CI table is read back in after the call completes.
SCOBI::SCRAPI(smoltData = STHD.trap, passageData = STHD.FPC,
              Primary = "GenStock", dateFormat = "%m/%d/%Y",
              Run = "fixed_", B = 2000)

fixed_ci <- read.csv("fixed_CIs.csv", skip = 1, header = TRUE,
                      row.names = 1, check.names = FALSE, fill = TRUE)
fixed_ci <- fixed_ci[!grepl("SampleSize", rownames(fixed_ci)), ]
```

```{r scrapi2-model, echo=TRUE}
# SCRAPI2 with ge_draws lets each bootstrap iteration pull one column from
# the daily GE posterior, so GE uncertainty flows into the abundance CI.
result_model <- smoltEASE::SCRAPI2(
  passageData = STHD.FPC, smoltData = STHD.trap, Primary = "GenStock",
  strat = "Week", dat = "SampleEndDate", tally = "SampleCount",
  samrate = "SampleRate", guidance = "GuidanceEfficiency", collaps = "Collapse",
  RTYPE = "W", alph = 0.05, dateFormat = "%m/%d/%Y", B = 2000,
  geDraws = ge_draws, Run = "model_"
)

```

```{r abundance-table, echo=FALSE}
abundance_tbl <- dplyr::bind_rows(
  as.data.frame(fixed_ci) |> tibble::rownames_to_column("Stock") |>
    dplyr::mutate(GE = "Current (NOAA fixed, SCOBI)"),
  as.data.frame(result_model$CI) |> tibble::rownames_to_column("Stock") |>
    dplyr::mutate(GE = "Model (spill-informed)")
) |>
  dplyr::filter(Stock == "WildSmolts") |>
  dplyr::mutate(`CI half-width (%)` = round((UCI - LCI) / (2 * Estimate) * 100, 1)) |>
  dplyr::select(GE, Estimate, LCI, UCI, `CI half-width (%)`)

knitr::kable(abundance_tbl,
             caption = "Wild steelhead smolt abundance, MY2025: current fixed-GE approach vs the spill-informed model.",
             format.args = list(big.mark = ","), digits = 0)
```

```{r abundance-plot, fig.cap="Wild steelhead smolt abundance under current fixed GE vs the spill-informed model, MY2025."}
ggplot(abundance_tbl, aes(GE, Estimate, color = GE)) +
  geom_pointrange(aes(ymin = LCI, ymax = UCI), linewidth = 0.8, size = 0.6,
                   show.legend = FALSE) +
  scale_color_manual(values = c("Current (NOAA fixed, SCOBI)" = "black",
                                "Model (spill-informed)"      = "red")) +
  scale_y_continuous(labels = comma) +
  labs(x = NULL, y = "Estimated wild steelhead smolts",
       title = "Wild steelhead abundance, MY2025") +
  theme_bw(base_size = 14)+expand_limits(y=0)
```

The fixed-GE approach holds GE near its pre-ROD historical level through peak
spill, which caps how many trapped fish need to be expanded to reach the true
run. Because the model estimates GE closer to single digits during the same
weeks, the same trap catch expands to a substantially larger run: the
model-GE estimate of 292,335 wild steelhead smolts is about 80% higher than
the fixed-GE estimate of 162,483, an increase of roughly 129,850 fish. The
model-GE interval is also wider (17% CI half-width vs 6% for the fixed-GE
run), and the extra width leans toward the upper end: the trap expansion
divides by GE, so a small GE draw produces a large abundance jump while a
large GE draw produces only a small one. That asymmetry is expected, not a
modeling artifact. The two intervals do not overlap: even the low end of the
model-GE estimate (264,701) sits well above the high end of the fixed-GE
estimate (172,451), so this is not just a shift in the central estimate but a
difference the two approaches' own uncertainty bounds agree on.

## References

Hance, D. J., J. M. Plumb, R. W. Perry, and K. F. Tiffan. 2024. Back from
the brink: estimating daily and annual abundance of natural-origin salmon
smolts from 30-years of mixed-origin capture-recapture data. Fisheries
Research 278:107098. https://doi.org/10.1016/j.fishres.2024.107098

McCann, J., B. Chockley, E. Cooper, G. Scheer, R. Tessier, S. Haeseker,
B. Lessard, T. Copeland, J. Ebel, A. Storch, and D. Rawding. 2023.
Comparative Survival Study of PIT-tagged spring/summer/fall Chinook,
summer steelhead, and Sockeye. 2023 Annual Report. BPA Contract No.
19960200. Available: https://www.fpc.org/documents/CSS/CSS%20Annual%20Report%202023.pdf.

Sandford, B. P., and S. G. Smith. 2002. Estimation of smolt-to-adult
return percentages for Snake River Basin anadromous salmonids, 1990-1997.
Journal of Agricultural, Biological, and Environmental Statistics
7(2):243-263. https://doi.org/10.1198/10857110260141274

*Data and software*

DART (Data Access in Real Time). Columbia Basin Research, University of
Washington School of Aquatic and Fishery Sciences. Available:
https://www.cbr.washington.edu/dart.

PTAGIS (Columbia Basin PIT Tag Information System). Pacific States Marine
Fisheries Commission. Available: https://www.ptagis.org.

Smith, D. `smoltEASE`: R package for guidance efficiency and smolt
abundance estimation. Available: https://github.com/dsmi313/smoltEASE.
