---
title: "GE Estimation at Lower Granite Dam Under Post-ROD Spill Conditions"
subtitle: "LGR juvenile abundance — a multistate GE model, and open questions"
author: "David Smith, IDFG Wild Salmon and Steelhead Program"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    fig_width: 10
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(SCOBI)
library(smoltEASE)
library(tidyverse)
library(lubridate)
library(scales)
library(knitr)
library(coda)
library(ISOweek)

input_dir  <- "C:/Users/david.smith/OneDrive - State of Idaho/Desktop/SCRAPI/SCRAPI2/Tim Meet"
output_dir <- input_dir
B          <- 2000
```

## Purpose

Dr. Hance — I work on wild steelhead and Chinook smolt abundance at LGR for
IDFG, and I am writing because the GE problem below led me straight to your
multistate PIT-tag work (Hance et al. 2020, 2024). I have built a guidance-
efficiency model closely modeled on that framework and folded its posterior into
the SCRAPI juvenile-abundance expansion (`smoltEASE::SCRAPI2`). I have not
contacted you before; I would simply value your read, as the originator of this
approach, on whether what I have done is defensible. Below I (1) restate why the
current GE inputs fail under post-ROD spill, (2) describe the model as built,
(3) show what it produces for MY2025 wild steelhead, and (4) lay out the specific
identifiability and structural questions I am least sure about. The model runs
and is internally coherent; I am not confident it is *right*, and the open
questions are real.

---

## Background

SCRAPI estimates wild steelhead smolt abundance at LGR by expanding the daily
juvenile bypass system (JBS) trap sample using daily guidance efficiency (GE) —
the fraction of outmigrating fish that enter the JBS and become available for
sampling. GE is currently provided by NOAA Fisheries NWFSC from the Sandford and
Smith (2002) mark-recapture framework, developed under pre-ROD, low-spill
conditions.

Under the 2020 Record of Decision, LGR now operates at 80–85% spill during peak
steelhead migration (April–May). A PIT antenna array at spillbay 1 (GRS, one of
eight spillbays), operational since 2020, provides direct daily counts of fish
entering the monitored spillbay. This array has never been incorporated into the
SCRAPI juvenile framework and its detection probability has never been formally
estimated for steelhead or spring–summer Chinook. The model below is a first
attempt to do exactly that. Current analyses focus on wild steelhead (MY2025);
the same framework will be applied to yearling spring–summer Chinook once a
defensible GE estimate is in hand.

### SCRAPI and guidance efficiency

SCRAPI estimates total passage by expanding the daily JBS trap sample:

$$N_i = \frac{c_i}{r_i \cdot \widehat{p}_i}, \qquad \hat{N} = \sum_{i=1}^{S} N_i$$

$c_i$ is the trap count, $r_i$ the daily trap sampling rate, and $\widehat{p}_i$
the estimated GE on day $i$. GE is not constant — it is strongly negatively
correlated with spill proportion, because spillway fish bypass the JBS entirely.
That correlation, under 80–85% spill, is the central issue: the expansion is
$1/\text{GE}$, so a small absolute error in a near-zero GE is an enormous error
in abundance.

### Why the existing GE inputs fail at high spill

**Sandford and Smith (2002)** estimate daily bypass detection probability as a
function of spill from GRJ/GOJ PIT detections, but were fit under low-spill
conditions. At 80–85% spill the regression is extrapolating well outside its
training range.

**McCann et al. (2023, Table 8.3)** provide a spill-dependent bypass-passage
model fit on MY 1998–2023 (logit scale):

$$\text{logit}(\hat{p}_i) = \beta_0 + \beta_1\,\text{stFlow}_i + \beta_2\,\text{PropSpill}_i + \beta_3\,\text{Weir}_i$$

with $\beta_0=0.85$ (so $\text{FGE}=\text{logit}^{-1}(0.85)\approx0.70$),
$\beta_1=0.01066$, $\beta_2=-0.08227$, $\beta_3=-0.34165$. It includes high-spill
years and so extrapolates better than Sandford–Smith, but it was not built to
supply GE to SCRAPI and, as shown below, **collapses to near-zero GE (≤0.5%) at
>75% spill** — which, propagated through $1/\text{GE}$, yields SCRAPI abundance
in the millions. It is useful as a directional check, not a usable input.

---

## Direct PIT-based GE, and the GRS detection problem

If we knew the GRS detection probability $\psi_d$ (the chance a spillway fish is
detected at bay 1 on day $d$), GE would follow directly from the daily GRJ/GRS
counts:

$$\widehat{\text{GE}}_d = 1 - \frac{n_{\text{GRS},d}/\psi_d}{n_{\text{GRS},d}/\psi_d + n_{\text{GRJ},d}} = \frac{n_{\text{GRJ},d}}{n_{\text{GRJ},d} + n_{\text{GRS},d}/\psi_d}$$

But GRS monitors only bay 1 of 8, so its effective detection probability is

$$\psi_d = \psi^* \cdot q_d$$

— intrinsic antenna efficiency $\psi^*$ times the fraction of spillway fish using
bay 1, $q_d$. The bay-1 weir is more attractive to surface-oriented fish than the
other bays, and that attraction varies with flow and operations, so $q_d$ is not
the hydraulic spill fraction. **Estimating $\psi_d$ is the crux of the whole
problem.** As a first look I bracket it at $\psi_d=0.50$ and $0.25$ below — not
estimates, just bounds to show SCRAPI's sensitivity. The model in the next
section replaces those assumptions with an *estimated* $\psi_d$.

---

## A multistate GE model that estimates $\psi_d$

This is the core of what I have built, implemented in
`smoltEASE::fit_ge_model()`. It is a stratum-level multistate mark–recapture
model on the first-detection histories of PIT-tagged fish at three arrays:
**GRJ** (LGR bypass), **GRS** (LGR spillbay 1), and **GOJ** (Little Goose
bypass). It jointly estimates, per stratum $s$:

- $\text{GE}_s$ — P(fish takes the JBS/bypass route);
- $\psi_s$ — P(spillway fish detected at GRS) — the bay-1 detection probability;
- $\phi_s$ — P(fish detected downstream at GOJ).

Each detected fish falls into one of five histories, with probabilities
conditioned on being detected somewhere:

$$
\begin{aligned}
(\text{GRJ, no GOJ}) &: \text{GE}_s(1-\phi_s) & (\text{GRJ, GOJ}) &: \text{GE}_s\phi_s\\
(\text{GRS, no GOJ}) &: (1-\text{GE}_s)\psi_s(1-\phi_s) & (\text{GRS, GOJ}) &: (1-\text{GE}_s)\psi_s\phi_s\\
(\text{no LGR, GOJ}) &: (1-\text{GE}_s)(1-\psi_s)\phi_s
\end{aligned}
$$

The **fifth cell — missed at both LGR arrays but recovered at GOJ — is what
identifies $\psi_s$** (and therefore corrects GE for the spillway fish GRS never
saw). This is exactly the two-route detection structure from Hance et al. (2024):
a routing probability (GE) and a spillway detection probability ($\psi_s$)
conditional on not taking the bypass, anchored by a downstream array.

Each parameter is a hierarchical logistic regression on spill (GE and $\psi$ on
LGR spill, $\phi$ on LGS spill), with weakly-informative priors on the GE and
$\psi$ coefficients and **informative priors on the $\phi$ (GOJ-detection)
coefficients taken from McCann et al. (2023, Table 8.3)** — because the
downstream detection probability is the least identifiable piece (see open
questions). Strata are defined at **weekly** resolution, with each week's GE
shrunk toward its **collapse-stratum mean** — a hyperparameter ($\mu_{\text{strat}}$)
estimated from the PIT detection histories, *not* the NOAA GE; the collapse
strata only define which weeks are grouped (the same week→stratum map used
operationally). This partial pooling lets sparse late-season weeks borrow
strength from their stratum group instead of producing high-leverage near-zero
GE. GE is then resolved to a **daily** value via the fitted spill regression. The
full JAGS specification is in the `smoltEASE` package (`fit_ge_model`).

```{r model-code, echo=TRUE, eval=FALSE}
# The model pipeline (smoltEASE). Heavy JAGS fit is cached to RDS below.
sthd_weekly <- transform(sthd_strata, Collapse = Week)        # one stratum per week
ge_data <- prep_ge_data(pit_raw, sthd_weekly, lgr_spill,
                        lgs_spill_data = lgs_spill, species = "sthd",
                        goj_site = "GOJ", parent_strata = sthd_strata)
ge_fit  <- fit_ge_model(ge_data, max_n = 500, seed = 42)      # nested weekly model
ge_draws <- generate_ge_draws(ge_fit, sthd_weekly,
              pass_dates = as.Date(STHD.FPC$SampleEndDate, "%m/%d/%Y"),
              B = B, daily_spill = lgr_spill)                 # daily GE posterior draws
```

---

## Data and fit

```{r load-data}
STHD.trap <- read.csv(file.path(input_dir, "MY2025STHD.trapData_formatted.csv"))
STHD.FPC  <- read.csv(file.path(input_dir, "MY2025STHDsmolts_FPCdata_strata_added.csv"))
STHD.FPC$SampleEndDate   <- format(as.Date(STHD.FPC$SampleEndDate,   "%m/%d/%Y"), "%m/%d/%Y")
STHD.trap$CollectionDate <- format(as.Date(STHD.trap$CollectionDate, "%Y-%m-%d"), "%m/%d/%Y")

# GSI posterior draws + genotyped trap subset (for the SCRAPI2 model run)
gsi_draws     <- read.delim(file.path(input_dir, "OmyLGRJ25C_gsiDraws_final.txt"),
                            sep = "\t", header = TRUE, check.names = FALSE)
STHD.trap.gsi <- STHD.trap[STHD.trap$MasterID %in% gsi_draws[[1]], ]

# Direct PIT GRJ/GRS counts by week (for the psi = 0.25 / 0.50 brackets)
pit <- read_csv(file.path(input_dir, "MY25_STHD.csv")) |>
  mutate(site = str_extract(`Site Name`, "^[A-Z]+"),
         obs_date = as_date(parse_date_time(`Obs Time Value`,
                            orders = c("mdy HM", "mdy HMS", "mdy HMSp")))) |>
  filter(site %in% c("GRJ", "GRS"),
         obs_date >= ymd("2025-03-01"), obs_date < ymd("2025-08-01")) |>
  group_by(`Tag Code`, site) |> slice_min(obs_date, with_ties = FALSE) |> ungroup() |>
  mutate(Week = isoweek(obs_date)) |>
  count(Week, site) |>
  pivot_wider(names_from = site, values_from = n, values_fill = 0) |>
  rename(n_grj = GRJ, n_grs = GRS) |>
  mutate(GE_pit_d50 = replace_na(n_grj / (n_grj + n_grs / 0.50), 0),
         GE_pit_d25 = replace_na(n_grj / (n_grj + n_grs / 0.25), 0))

# Spill (DART) for the McCann curve and the weekly spill bars
dart <- read_csv(file.path(input_dir, "DART Spill.csv")) |>
  mutate(date = mdy(date),
         spill_per    = as.numeric(spill_kcfs) / as.numeric(outflow_kcfs) * 100,
         outflow_kcfs = as.numeric(outflow_kcfs)) |>
  filter(date >= ymd("2025-03-01"), date < ymd("2025-08-01")) |>
  mutate(Week = isoweek(date)) |>
  group_by(Week) |>
  summarise(spill_per = replace_na(mean(spill_per, na.rm = TRUE), 0),
            outflow_kcfs = mean(outflow_kcfs, na.rm = TRUE),
            week_start = min(date), .groups = "drop")
```

```{r model-inputs}
# Inputs for the multistate model: GRJ/GRS/GOJ first-detections + daily spill
pit_raw <- read_csv(file.path(input_dir, "MY25_STHD.csv")) |>
  mutate(site = str_extract(`Site Name`, "^[A-Z]+"),
         obs_date = as_date(parse_date_time(`Obs Time Value`,
                            orders = c("mdy HM", "mdy HMS", "mdy HMSp")))) |>
  filter(`Rear Type Code` %in% c("W", "H", "U"),
         site %in% c("GRJ", "GRS", "GOJ"),
         obs_date >= ymd("2025-03-24"), obs_date < ymd("2025-08-01")) |>
  group_by(`Tag Code`, site) |> slice_min(obs_date, with_ties = FALSE) |> ungroup() |>
  rename(tag = `Tag Code`, det_date = obs_date)

lgr_spill <- read_csv(file.path(input_dir, "DART Spill.csv")) |>
  mutate(Date = mdy(date),
         spill.per = replace_na(as.numeric(spill_kcfs)/as.numeric(outflow_kcfs)*100, 0)) |>
  filter(Date >= ymd("2025-03-24"), Date < ymd("2025-08-01")) |>
  select(Date, spill.per)

lgs_spill <- bind_rows(read_csv(file.path(input_dir, "spill.all.csv")),
                       read_csv(file.path(input_dir, "spill.all2.csv")),
                       read_csv(file.path(input_dir, "spill.all3.csv"))) |>
  mutate(Date = as_date(mdy_hm(DateTime)), spill.per = HourlySpill / HourlyFlow * 100) |>
  filter(Site == "LGS", Date >= ymd("2025-03-24"), Date < ymd("2025-08-01")) |>
  group_by(Date) |> summarise(spill.per = mean(spill.per, na.rm = TRUE), .groups = "drop")

strata_raw  <- readxl::read_xlsx(file.path(input_dir, "Smolt Strata.xlsx"), sheet = 1)
sthd_start  <- which(strata_raw[[1]] == "WeekNumber" & strata_raw[[2]] == "STHD_n")
sthd_strata <- strata_raw[(sthd_start + 1):nrow(strata_raw), c(1, 3)] |>
  set_names(c("Week", "Collapse")) |> filter(!is.na(Week)) |>
  mutate(Week = as.integer(Week), Collapse = as.integer(Collapse))
```

```{r fit-model}
# Fit the weekly+nested model once and cache (memo-specific file; re-knits load it).
rds <- file.path(output_dir, "ge_fit_weekly_MY2025_STHD.rds")
if (file.exists(rds)) {
  obj      <- readRDS(rds); ge_fit <- obj$ge_fit; ge_draws <- obj$ge_draws
} else {
  sthd_weekly <- transform(sthd_strata, Collapse = Week)
  ge_data <- prep_ge_data(pit_raw, sthd_weekly, lgr_spill, lgs_spill_data = lgs_spill,
                          species = "sthd", goj_site = "GOJ", parent_strata = sthd_strata)
  ge_fit  <- fit_ge_model(ge_data, max_n = 500, seed = 42)
  ge_draws <- generate_ge_draws(ge_fit, sthd_weekly,
                pass_dates = as.Date(STHD.FPC$SampleEndDate, "%m/%d/%Y"),
                B = B, daily_spill = lgr_spill)
  saveRDS(list(ge_fit = ge_fit, ge_draws = ge_draws), rds)
}

# Weekly model GE (= psi posterior) and the estimated GRS detection prob (p)
vn   <- coda::varnames(ge_fit$samples)
smat <- as.matrix(ge_fit$samples)
ge_w <- smat[, grep("^psi\\[", vn), drop = FALSE]   # GE (bypass route prob)
p_w  <- smat[, grep("^p\\[",   vn), drop = FALSE]   # GRS detection prob (psi_d in the prose)

model_ge <- tibble(
  Week     = ge_fit$obs_strata$stratum,
  GE_model = colMeans(ge_w),
  GE_lo    = apply(ge_w, 2, quantile, 0.025),
  GE_hi    = apply(ge_w, 2, quantile, 0.975),
  psi_det  = colMeans(p_w),
  psi_lo   = apply(p_w, 2, quantile, 0.025),
  psi_hi   = apply(p_w, 2, quantile, 0.975)
)
```

---

## Guidance efficiency comparison, MY2025

```{r ge-plot, fig.cap="Weekly GE for MY2025 wild steelhead. Black: current NOAA GE. Green: McCann et al. (2023) logit (comparison only). Blue/pink: direct PIT-based GE assuming 50% / 25% GRS detection. Red band: the multistate model GE (posterior mean + 95% CI). Bars: weekly mean spill proportion."}
wk_date <- function(w) ISOweek2date(paste0("2025-W", str_pad(w, 2, pad = "0"), "-1"))

# Model GE by ISO week, built from the DAILY posterior draws (keyed to real
# passage dates) so it lands on the correct calendar weeks regardless of how the
# fit was stratified.
draws_mat <- as.matrix(ge_draws[, -1])
draw_wk   <- isoweek(as.Date(ge_draws$SampleEndDate, "%m/%d/%Y"))
model_wk  <- purrr::map_dfr(sort(unique(draw_wk)), function(w) {
  m <- colMeans(draws_mat[draw_wk == w, , drop = FALSE])   # week-mean GE per draw
  tibble(Week = w, GE_model = mean(m),
         GE_lo = quantile(m, 0.025), GE_hi = quantile(m, 0.975))
})

# McCann curve
css_ge <- dart |>
  mutate(GE_css = plogis(0.85 + 0.01066*(outflow_kcfs - 24) - 0.08227*spill_per - 0.34165*1)) |>
  select(Week, GE_css)

ge_noaa <- STHD.FPC |> group_by(Week) |>
  summarise(GE_noaa = mean(GuidanceEfficiency, na.rm = TRUE), .groups = "drop")

ge_lines <- ge_noaa |>
  left_join(css_ge, by = "Week") |>
  left_join(pit |> select(Week, GE_pit_d50, GE_pit_d25), by = "Week") |>
  left_join(model_wk |> select(Week, GE_model), by = "Week") |>
  mutate(week_start = wk_date(Week)) |>
  pivot_longer(c(GE_noaa, GE_css, GE_pit_d50, GE_pit_d25, GE_model),
               names_to = "method", values_to = "GE") |>
  mutate(method = recode(method,
    GE_noaa = "Current (NOAA) GE", GE_css = "McCann et al. (2023) logit",
    GE_pit_d50 = "PIT GE (det = 0.50)", GE_pit_d25 = "PIT GE (det = 0.25)",
    GE_model = "Multistate model GE"))

ribbon <- model_wk |> mutate(week_start = wk_date(Week))

# x-axis fits JUST our GE-estimate window (the model / passage weeks), not the
# wider McCann curve or spill series, which are clipped to it.
xr   <- range(wk_date(model_wk$Week))
bars <- dart |> mutate(week_start = wk_date(Week)) |>
  filter(week_start >= xr[1], week_start <= xr[2])

pal <- c("Current (NOAA) GE" = "black", "McCann et al. (2023) logit" = "#1a9641",
         "PIT GE (det = 0.50)" = "#2166ac", "PIT GE (det = 0.25)" = "#d01c8b",
         "Multistate model GE" = "#d73027")

ggplot() +
  geom_col(data = bars, aes(week_start, spill_per/100), fill = "steelblue",
           alpha = 0.18, width = 5) +
  geom_ribbon(data = ribbon, aes(week_start, ymin = GE_lo, ymax = GE_hi),
              fill = "#d73027", alpha = 0.15) +
  geom_line(data = ge_lines, aes(week_start, GE, color = method), linewidth = 0.9, na.rm = TRUE) +
  geom_point(data = ge_lines, aes(week_start, GE, color = method), size = 1.8, na.rm = TRUE) +
  scale_color_manual(values = pal) +
  scale_y_continuous("Guidance efficiency", labels = percent_format(accuracy = 1),
                     limits = c(0, 1),
                     sec.axis = sec_axis(~ ., name = "Spill proportion",
                                         labels = percent_format(accuracy = 1))) +
  coord_cartesian(xlim = xr) +
  labs(x = NULL, color = NULL,
       title = "Weekly guidance efficiency — MY2025 wild steelhead",
       subtitle = "Five GE methods; bars show weekly mean spill") +
  theme_bw(base_size = 12) + theme(legend.position = "bottom")
```

During pre-spill weeks (13–14) all methods converge near 40–60% GE. Once high
spill begins (week 15) the NOAA GE stabilizes near 13% through the peak migration
(weeks 15–22). The McCann logit collapses to ≤0.5% — biologically implausible
when expanded. The PIT brackets give 3–10% (det = 0.50) and 2–5% (det = 0.25).
**The multistate model GE falls within those brackets through the high-spill
period**, with credible intervals that widen exactly where the data thin
(late-season weeks). Its estimated GRS detection probability $\psi_d$ — the thing
I previously had to assume — is summarized below:

```{r psi-table}
model_ge |>
  transmute(Week,
            `Model GE` = percent(GE_model, 0.1),
            `GE 95% CI` = paste0(percent(GE_lo, 0.1), " – ", percent(GE_hi, 0.1)),
            `GRS detection psi_d` = percent(psi_det, 0.1),
            `psi_d 95% CI` = paste0(percent(psi_lo, 0.1), " – ", percent(psi_hi, 0.1))) |>
  kable(align = "lrrrr",
        caption = "Model-estimated weekly GE and GRS bay-1 detection probability (psi_d).")
```

---

## Abundance: NOAA fixed GE vs the model

`smoltEASE::SCRAPI2` expands the trap exactly as SCRAPI does, but consumes the GE
posterior draws (one per bootstrap iteration), so GE uncertainty propagates into
the CIs. With degenerate draws it reproduces `SCOBI::SCRAPI` (validated
separately). Here I compare the current NOAA fixed GE against the model GE.

```{r run-scrapi, echo=TRUE, eval=FALSE}
# Current (NOAA) GE: GuidanceEfficiency column already in the FPC
res_noaa <- SCRAPI2(smoltData = STHD.trap, passageData = STHD.FPC, Primary = "GenStock",
  strat = "Week", dat = "SampleEndDate", tally = "SampleCount", samrate = "SampleRate",
  guidance = "GuidanceEfficiency", collaps = "Collapse", RTYPE = "W", alph = 0.05,
  dateFormat = "%m/%d/%Y", Run = "MY2025_wSthd_GE_NOAA", B = B)

# Multistate model GE (posterior draws) + GSI uncertainty
res_model <- SCRAPI2(smoltData = STHD.trap.gsi, passageData = STHD.FPC, Primary = "GenStock",
  Secondary = "fwAge", geDraws = ge_draws, gsiDraws = gsi_draws, fishID = "MasterID",
  strat = "Week", dat = "SampleEndDate", tally = "SampleCount", samrate = "SampleRate",
  guidance = "GuidanceEfficiency", collaps = "Collapse", RTYPE = "W", alph = 0.05,
  dateFormat = "%m/%d/%Y", Run = "MY2025_wSthd_SCRAPI2_BayesGE", B = B)
```

```{r scrapi-results}
ci_files <- c(NOAA  = "MY2025_wSthd_GE_NOAACIs.csv",
              Model = "MY2025_wSthd_SCRAPI2_BayesGECIs.csv")
ready <- all(file.exists(file.path(output_dir, ci_files)))

read_total <- function(filename, label) {
  lines <- readLines(file.path(output_dir, filename))
  dl    <- lines[!grepl("^(RearSampleSize|PrimeSampleSize|SecondSampleSize|W -)", lines) & lines != ""]
  df    <- read.csv(text = paste(dl, collapse = "\n")); names(df)[1] <- "Group"
  tot   <- df |> filter(Group == "WildSmolts")
  tibble(method = label, N = as.numeric(tot$Estimate),
         lower = as.numeric(tot$LCI), upper = as.numeric(tot$UCI))
}

if (ready) {
  results <- bind_rows(
    read_total(ci_files["NOAA"],  "Current (NOAA) GE"),
    read_total(ci_files["Model"], "Multistate model GE + GSI")
  ) |> mutate(method = fct_inorder(method))
}
```

```{r results-table}
if (ready) {
  results |>
    mutate(N = comma(N), CI = paste0(comma(lower), " – ", comma(upper))) |>
    select(`GE method` = method, `Wild steelhead estimate` = N, `95% CI` = CI) |>
    kable(align = "lrr",
          caption = "SCRAPI/SCRAPI2 wild steelhead abundance — NOAA fixed GE vs multistate model GE (with full GE + GSI uncertainty).")
} else {
  cat("_Run the `run-scrapi` chunk to populate the abundance comparison._")
}
```

```{r scrapi-plot, fig.width=7, fig.height=5, eval=FALSE}
ggplot(results, aes(method, N, ymin = lower, ymax = upper, color = method)) +
  geom_pointrange(size = 0.8, linewidth = 1.2) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  scale_color_manual(values = c("Current (NOAA) GE" = "black",
                                "Multistate model GE + GSI" = "#d73027")) +
  labs(x = NULL, y = "Wild steelhead abundance",
       title = "SCRAPI2 wild steelhead abundance — MY2025",
       subtitle = "NOAA fixed GE vs multistate model GE with full GE + GSI uncertainty") +
  theme_bw(base_size = 13) + theme(legend.position = "none")
```

Lower GE means a larger $1/\text{GE}$ expansion, so the model — which sits below
the NOAA ~13% through peak spill — produces a higher wild-steelhead estimate.
This has direct SAR consequences: an underestimated juvenile abundance inflates
SAR, with regulatory implications under BiOp performance standards. The model's
contribution beyond the point estimate is **honest uncertainty**: its CIs widen
in exactly the weeks where GE is least constrained.

---

## Open questions — where I would value your read

The model runs and is internally coherent, but several choices are load-bearing
and I am not sure they are right. These are the points I would most like your
thoughts on.

1. **Identifiability of $\psi_d$ from GOJ alone.** The GRS detection probability
   is identified only by fish missed at *both* LGR arrays and recovered at GOJ
   (the fifth detection-history cell). Under 80–85% spill, a fish that uses the
   LGR spillway and is missed by GRS will very likely use the spillway again at
   LGS and never reach the GOJ bypass — the route-affinity problem inherent to
   this geometry. That shrinks the identifying cell toward zero. **(a)** Is GOJ recovery
   sufficient, or do we need to bring in LMJ/MCJ/IHR (with their own spill-
   dependent bypass probabilities)? **(b)** Does the McCann-informed prior on the
   GOJ-detection $\phi$ genuinely stabilize this, or is it papering over weak
   identification — i.e., is the posterior on $\psi_d$ mostly prior?

2. **Modeling $q_d$ (bay-1 attraction).** I currently absorb
   $\psi_d = \psi^* q_d$ into a single spill-dependent detection probability,
   rather than separating intrinsic antenna efficiency from the fraction of
   spillway fish attracted to bay 1. DART's `percent_spill_spillbay1` is a
   candidate covariate for $q_d$. Worth adding, or does the lumped spill
   regression capture enough given the weir's flow-dependent attraction?

3. **Overdispersion / multinomial sample size.** I cap each stratum's effective
   multinomial sample size (`max_n = 500`, McCann-style data thinning) so that
   data-rich weeks with tens of thousands of detections do not collapse the GE
   posterior to an implausibly tight point. Would you prefer an explicit
   overdispersion model (Dirichlet-multinomial, or a per-observation random
   effect) that lets the data set its own effective $n$?

4. **Separate GE then SCRAPI, vs joint estimation.** I estimate GE in a
   standalone multistate model and plug the posterior draws into SCRAPI's
   $1/\text{GE}$ expansion. Your 2024 TSCJS estimates abundance jointly. Is the
   two-step decoupling defensible, or am I losing coherence / understating
   uncertainty by not propagating the GE–abundance covariance?

5. **Resolution and pooling.** GE is fit weekly, nested in the collapse strata
   (partial pooling toward a *data-estimated* stratum mean — the collapse strata
   are only the grouping, not NOAA GE), then resolved to daily via the spill
   regression. For the sparse late-season weeks this leans heavily on the
   regression. Is weekly-nested the right granularity, or would you fit daily
   with a temporal smoother?

---

## Summary

Under post-ROD operations, GE at LGR during peak migration is most plausibly in
the **2–10% range, not the ~13%** currently used by SCRAPI. The multistate model
— estimating the GRS bay-1 detection probability jointly with GE from GRJ/GRS/GOJ
detection histories, hierarchically across weeks, in the Hance et al. (2020,
2024) two-route structure — gives GE estimates inside the 25–50% GRS-detection
brackets, with uncertainty that grows where the data thin, and propagates that
uncertainty into SCRAPI abundance and ultimately SAR.

The model is usable, but its credibility rests on the identifiability of $\psi_d$
under high spill, where the downstream recoveries that should pin it are
themselves suppressed by re-spill — the exact problem your framework was built to
handle. Your read on questions 1 and 4 in particular would shape whether we keep
the two-step GE→SCRAPI approach or move toward a joint abundance model.

---

*Analysis in R using PTAGIS interrogation data, DART daily operations records,
the `SCOBI`/`smoltEASE` packages, and McCann et al. (2023) Table 8.3
coefficients. GE model: `smoltEASE::fit_ge_model` (multistate, hierarchical,
JAGS).*
