---
title: "A multistate guidance-efficiency model for Lower Granite juvenile abundance"
subtitle: "Model description — GE estimation under post-ROD spill"
author: "David Smith, IDFG Wild Salmon and Steelhead Program"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    fig_width: 10
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(tidyverse)
library(lubridate)
library(scales)
library(knitr)
library(ISOweek)

input_dir <- "C:/Users/david.smith/OneDrive - State of Idaho/Desktop/SCRAPI/SCRAPI2/Tim Meet"
```

This note documents a guidance-efficiency (GE) model for Lower Granite Dam (LGR)
juvenile abundance and is circulated for comment. It is **methods only** — it
describes the problem, the multistate model (including the complete JAGS
specification), and the equation that converts the model's stratum-level
estimates into daily GE. No abundance results are shown.

---

## The GE problem

SCRAPI estimates total juvenile passage at LGR by expanding the daily juvenile
bypass system (JBS) trap sample by the daily guidance efficiency — the fraction
of outmigrants that enter the JBS:

$$N_i = \frac{c_i}{r_i \cdot \widehat{GE}_i}, \qquad \hat{N} = \sum_{i=1}^{S} N_i$$

$c_i$ is the trap count, $r_i$ the trap sampling rate, and $\widehat{GE}_i$ the
GE on day $i$. Because the expansion is $1/GE$, GE is the most consequential and
the most uncertain input. GE is strongly **negatively** correlated with spill —
spillway fish bypass the JBS entirely — and under the 2020 Record of Decision LGR
now runs at **80–85% spill** through the April–May steelhead peak.

The GE values SCRAPI currently uses come from NOAA NWFSC via the Sandford and
Smith (2002) mark–recapture framework, developed under **pre-ROD, low-spill**
conditions. At 80–85% spill that framework is extrapolating well outside the
range it was fit on.

---

## Current GE is likely too high

The figure below compares, by week for MY2025 wild steelhead, the current NOAA
GE against (i) the McCann et al. (2023, Table 8.3) spill-dependent bypass-passage
logit model, and (ii) a direct PIT-based GE computed from GRJ (bypass) and GRS
(spillbay-1) detection counts under two assumed GRS detection probabilities
(0.50, 0.25; see next section). It is shown only to motivate the model: through
the high-spill period the NOAA GE sits near 13%, well above every other line.

```{r load-data}
# NOAA weekly GE (already merged into the FPC passage file)
STHD.FPC <- read.csv(file.path(input_dir, "MY2025STHDsmolts_FPCdata_strata_added.csv"))
STHD.FPC$SampleEndDate <- format(as.Date(STHD.FPC$SampleEndDate, "%m/%d/%Y"), "%m/%d/%Y")

# Direct PIT GRJ/GRS first-detection counts by ISO week
pit <- read_csv(file.path(input_dir, "MY25_STHD.csv")) |>
  mutate(site = str_extract(`Site Name`, "^[A-Z]+"),
         obs_date = as_date(parse_date_time(`Obs Time Value`,
                            orders = c("mdy HM", "mdy HMS", "mdy HMSp")))) |>
  filter(site %in% c("GRJ", "GRS"),
         obs_date >= ymd("2025-03-01"), obs_date < ymd("2025-08-01")) |>
  group_by(`Tag Code`, site) |> slice_min(obs_date, with_ties = FALSE) |> ungroup() |>
  mutate(Week = isoweek(obs_date)) |> count(Week, site) |>
  pivot_wider(names_from = site, values_from = n, values_fill = 0) |>
  rename(n_grj = GRJ, n_grs = GRS) |>
  mutate(GE_pit_d50 = replace_na(n_grj / (n_grj + n_grs / 0.50), 0),
         GE_pit_d25 = replace_na(n_grj / (n_grj + n_grs / 0.25), 0))

# Weekly spill (DART) for the McCann curve and the spill bars
dart <- read_csv(file.path(input_dir, "DART Spill.csv")) |>
  mutate(date = mdy(date),
         spill_per = as.numeric(spill_kcfs) / as.numeric(outflow_kcfs) * 100,
         outflow_kcfs = as.numeric(outflow_kcfs)) |>
  filter(date >= ymd("2025-03-01"), date < ymd("2025-08-01")) |>
  mutate(Week = isoweek(date)) |>
  group_by(Week) |>
  summarise(spill_per = replace_na(mean(spill_per, na.rm = TRUE), 0),
            outflow_kcfs = mean(outflow_kcfs, na.rm = TRUE), .groups = "drop")
```

```{r ge-plot, fig.cap="Weekly GE for MY2025 wild steelhead. Black: current NOAA GE. Green: McCann et al. (2023) logit (comparison only). Blue/pink: direct PIT-based GE assuming 50% / 25% GRS detection. Bars: weekly mean spill proportion. The multistate model's GE is intentionally not shown here."}
wk_date <- function(w) ISOweek2date(paste0("2025-W", str_pad(w, 2, pad = "0"), "-1"))

css_ge  <- dart |>   # McCann logit, Weir = 1 (SSW in operation)
  mutate(GE_css = plogis(0.85 + 0.01066*(outflow_kcfs - 24) - 0.08227*spill_per - 0.34165*1)) |>
  select(Week, GE_css)
ge_noaa <- STHD.FPC |> group_by(Week) |>
  summarise(GE_noaa = mean(GuidanceEfficiency, na.rm = TRUE), .groups = "drop")

ge_lines <- ge_noaa |>
  left_join(css_ge, by = "Week") |>
  left_join(pit |> select(Week, GE_pit_d50, GE_pit_d25), by = "Week") |>
  mutate(week_start = wk_date(Week)) |>
  pivot_longer(c(GE_noaa, GE_css, GE_pit_d50, GE_pit_d25),
               names_to = "method", values_to = "GE") |>
  mutate(method = recode(method,
    GE_noaa = "Current (NOAA) GE", GE_css = "McCann et al. (2023) logit",
    GE_pit_d50 = "PIT GE (det = 0.50)", GE_pit_d25 = "PIT GE (det = 0.25)"))

xr   <- range(ge_lines$week_start[!is.na(ge_lines$GE)])
bars <- dart |> mutate(week_start = wk_date(Week)) |>
  filter(week_start >= xr[1], week_start <= xr[2])

pal <- c("Current (NOAA) GE" = "black", "McCann et al. (2023) logit" = "#1a9641",
         "PIT GE (det = 0.50)" = "#2166ac", "PIT GE (det = 0.25)" = "#d01c8b")

ggplot() +
  geom_col(data = bars, aes(week_start, spill_per/100), fill = "steelblue",
           alpha = 0.18, width = 5) +
  geom_line(data = ge_lines, aes(week_start, GE, color = method), linewidth = 0.9, na.rm = TRUE) +
  geom_point(data = ge_lines, aes(week_start, GE, color = method), size = 1.8, na.rm = TRUE) +
  scale_color_manual(values = pal) +
  scale_y_continuous("Guidance efficiency", labels = percent_format(accuracy = 1),
                     limits = c(0, 1),
                     sec.axis = sec_axis(~ ., name = "Spill proportion",
                                         labels = percent_format(accuracy = 1))) +
  coord_cartesian(xlim = xr) +
  labs(x = NULL, color = NULL,
       title = "Weekly guidance efficiency — MY2025 wild steelhead",
       subtitle = "Current NOAA GE vs spill-based alternatives; bars show weekly mean spill") +
  theme_bw(base_size = 12) + theme(legend.position = "bottom")
```

In pre-spill weeks the methods agree (~40–60%). Once high spill begins the NOAA
GE stabilizes near 13% while the spill-based lines fall to single digits (McCann
collapses to ≤0.5%; the PIT brackets to 2–10%). The point is simply that the
current ~13% looks too high under post-ROD spill. The model below estimates GE
directly from the PIT detections rather than assuming the GRS detection
probability.

---

## Direct PIT GE and the GRS detection problem

If the GRS detection probability $\psi_d$ (the chance a spillway fish is detected
at the monitored bay-1 antenna on day $d$) were known, GE would follow from the
daily GRJ/GRS counts:

$$\widehat{GE}_d = \frac{n_{\text{GRJ},d}}{n_{\text{GRJ},d} + n_{\text{GRS},d}/\psi_d}$$

But GRS monitors only 1 of 8 spillbays, so its effective detection probability is

$$\psi_d = \psi^* \cdot q_d$$

— intrinsic antenna efficiency $\psi^*$ times the fraction of spillway fish using
bay 1, $q_d$ (the bay-1 weir is more attractive to surface-oriented fish, and that
attraction varies with flow). Neither $\psi^*$ nor $q_d$ is known, so the PIT
ratio above can only be bracketed. The multistate model **estimates** $\psi_d$ as
a parameter, using downstream recoveries at Little Goose for identification.

---

## The multistate model

The model is a stratum-level multistate mark–recapture model on the
first-detection histories of PIT-tagged fish at three arrays: **GRJ** (LGR
bypass), **GRS** (LGR spillbay 1), and **GOJ** (Little Goose bypass). Per stratum
$s$ it estimates three probabilities, which in the JAGS code (below) are named:

| Code | Quantity |
|---|---|
| `psi[s]`  | **GE** — P(fish takes the JBS / bypass route) |
| `p[s]`    | **GRS detection** — P(spillway fish detected at bay 1) — the $\psi_d$ above |
| `phi[s]`  | **GOJ detection** — P(fish detected downstream at Little Goose) |

Each detected fish falls into one of five histories. Conditioned on being
detected somewhere, the cell probabilities are

$$
\begin{aligned}
(\text{GRJ, no GOJ}) &: \psi(1-\phi) & (\text{GRJ, GOJ}) &: \psi\phi\\
(\text{GRS, no GOJ}) &: (1-\psi)\,p\,(1-\phi) & (\text{GRS, GOJ}) &: (1-\psi)\,p\,\phi\\
(\text{no LGR, GOJ}) &: (1-\psi)(1-p)\phi
\end{aligned}
$$

The fifth cell — missed at **both** LGR arrays but recovered at GOJ — is what
identifies the GRS detection probability `p` (and therefore corrects GE for the
spillway fish GRS never saw). Each parameter is a hierarchical logistic
regression on spill (`psi`, `p` on LGR spill; `phi` on LGS spill), and weeks are
**nested** in coarser collapse strata: each week's level shrinks toward a
data-estimated stratum mean (`mu_strat`), which in turn shrinks toward a global
mean. The pipeline is three calls in the `smoltEASE` package:

```{r pipeline, echo=TRUE, eval=FALSE}
sthd_weekly <- transform(sthd_strata, Collapse = Week)        # one stratum per week
ge_data <- prep_ge_data(pit_raw, sthd_weekly, lgr_spill,      # GRJ/GRS/GOJ histories + spill
                        lgs_spill_data = lgs_spill, species = "sthd",
                        goj_site = "GOJ", parent_strata = sthd_strata)
ge_fit  <- fit_ge_model(ge_data, max_n = 500, seed = 42)      # JAGS fit (below)
ge_draws <- generate_ge_draws(ge_fit, sthd_weekly,           # stratum -> daily GE draws
              pass_dates = passage_dates, B = 2000, daily_spill = lgr_spill)
```

### The JAGS model (complete)

This is the exact model `fit_ge_model()` fits (the nested form; the single-level
form drops `mu_strat`/`sigma_strat` and writes `logit_psi[s] ~ dnorm(alpha + beta
* lgr_spill_std[s], tau_psi)`). Every line is annotated.

```
model {

  eps       <- 1.0E-9                     # tiny floor so no cell prob is exactly 0
  tau_psi   <- pow(sigma_psi,   -2)       # precisions = 1/sigma^2  (JAGS dnorm uses precision)
  tau_strat <- pow(sigma_strat, -2)       #   sigma_psi   : week-to-week (within-stratum) SD of logit(GE)
  tau_p     <- pow(sigma_p,     -2)       #   sigma_strat : stratum-to-stratum SD of the GE level
  tau_phi   <- pow(sigma_phi,   -2)

  for (s in 1:S) {                        # s = stratum (here, one per week)

    # --- GE (psi): nested hierarchical logistic regression on LGR spill ---
    #   each week's logit(GE) is centered on its PARENT collapse-stratum mean
    #   (mu_strat) plus a spill slope; beta is the within-stratum spill effect.
    logit_psi[s] ~ dnorm(mu_strat[parent[s]] + beta * lgr_spill_std[s], tau_psi)
    psi[s] <- ilogit(logit_psi[s])        # psi[s] = GE for stratum s

    # --- GRS bay-1 detection probability (p): logistic regression on LGR spill ---
    logit_p[s] ~ dnorm(alpha_p + beta_p * lgr_spill_std[s], tau_p)
    p[s] <- ilogit(logit_p[s])            # p[s] = P(spillway fish seen at GRS) = psi_d

    # --- GOJ (Little Goose) detection probability (phi): on LGS spill ---
    logit_phi[s] ~ dnorm(alpha_phi + beta_phi * lgs_spill_std[s], tau_phi)
    phi[s] <- ilogit(logit_phi[s])        # phi[s] = P(detected downstream at GOJ)

    # --- five detection-history cell probabilities (unnormalised) ---
    pi_raw[s, 1] <- psi[s] * (1 - phi[s])               + eps   # GRJ, not GOJ  (bypass, missed downstream)
    pi_raw[s, 2] <- psi[s] * phi[s]                     + eps   # GRJ and GOJ   (bypass, seen downstream)
    pi_raw[s, 3] <- (1 - psi[s]) * p[s] * (1 - phi[s])  + eps   # GRS, not GOJ  (spillway, detected at bay 1)
    pi_raw[s, 4] <- (1 - psi[s]) * p[s] * phi[s]        + eps   # GRS and GOJ
    pi_raw[s, 5] <- (1 - psi[s]) * (1 - p[s]) * phi[s]  + eps   # no LGR, GOJ   (spillway, MISSED at bay 1,
                                                               #                recovered downstream) -> identifies p

    # --- condition on being detected somewhere (we only see fish that were seen) ---
    p_seen[s] <- pi_raw[s,1] + pi_raw[s,2] + pi_raw[s,3] +
                 pi_raw[s,4] + pi_raw[s,5]
    for (k in 1:5) {
      pi_obs[s, k] <- pi_raw[s, k] / p_seen[s]          # cell probs given detection
    }
  }

  # --- multinomial likelihood, ONLY for strata that have detection-history data ---
  #   lik_idx / S_lik index the strata with N_seen > 0; data-empty weeks contribute
  #   no likelihood term but still get a hierarchical psi (so they stay estimable).
  for (j in 1:S_lik) {
    n[lik_idx[j], 1:5] ~ dmulti(pi_obs[lik_idx[j], 1:5], N_seen[lik_idx[j]])
  }

  # --- nesting: collapse-stratum means shrink to a global mean (partial pooling) ---
  for (g in 1:n_strat) {
    mu_strat[g] ~ dnorm(alpha, tau_strat)
  }

  # --- priors: GE regression (alpha = global mean; beta = spill slope) ---
  alpha       ~ dt(0, pow(2, -2), 7)      # Student-t(scale 2, df 7): weakly informative
  beta        ~ dt(0, pow(1, -2), 7) T(, 0)  # T(,0): spill can only LOWER GE, never raise it
  sigma_psi   ~ dunif(0.05, 3)
  sigma_strat ~ dunif(0.05, 3)

  # --- priors: GRS detection regression ---
  alpha_p  ~ dt(0, pow(2, -2), 7)
  beta_p   ~ dt(0, pow(1, -2), 7)
  sigma_p  ~ dunif(0.05, 3)

  # --- priors: GOJ detection regression -- INFORMATIVE, from McCann et al. (2023)
  #     Table 8.3 (LGS steelhead bypass-passage logit coefficients). phi is the
  #     least identifiable piece, so it is anchored by these external estimates.
  alpha_phi ~ dnorm(-2.60, pow(0.30, -2))      # intercept ~ N(-2.60, sd 0.30)
  beta_phi  ~ dnorm(-1.78, pow(0.25, -2)) T(, 0)  # spill slope ~ N(-1.78, sd 0.25), <= 0
  sigma_phi ~ dunif(0.05, 1)
}
```

Two data-preparation choices worth flagging:

- **Multinomial thinning (`max_n = 500`).** Before fitting, each stratum's five
  history counts are scaled down so the effective multinomial size `N_seen` is at
  most `max_n`. The multinomial treats each detection as an independent replicate;
  with tens of thousands of detections it would otherwise drive the GE posterior
  to an implausibly tight point. This caps the per-stratum information (a
  data-thinning device used by McCann et al. 2023). It is a candidate for
  replacement by an explicit overdispersion model.
- **Empty strata.** A week with no detection-history fish contributes no
  likelihood term (`lik_idx` excludes it) but still receives a regression-based
  GE, so sparse weeks remain estimable by partial pooling rather than breaking.

---

## From weekly estimates to daily GE

`fit_ge_model()` estimates GE at the stratum (week) level. SCRAPI expands trap
counts **daily**, and GE moves with daily spill, so the weekly posterior is
resolved to a daily value using the same fitted spill regression. Spill is
standardised with the mean $\bar s$ and SD $s_{sd}$ used in the fit. For a draw of
stratum $s$ with GE $\psi_s$, the realised process deviation on the logit scale is
recovered, then the day's spill is substituted:

$$\text{std}_d = \frac{\text{spill}_d - \bar s}{s_{sd}}, \qquad
  \varepsilon_s = \operatorname{logit}(\psi_s) - \big(\alpha + \beta\,\text{std}_s\big),$$

$$\boxed{\;\widehat{GE}_d
   = \operatorname{logit}^{-1}\!\big(\alpha + \beta\,\text{std}_d + \varepsilon_s\big)
   = \operatorname{logit}^{-1}\!\big(\operatorname{logit}(\psi_s) + \beta\,(\text{std}_d - \text{std}_s)\big)\;}$$

So the stratum's fitted GE level $\psi_s$ is held fixed (through $\varepsilon_s$)
while the day's spill deviation from the stratum mean, scaled by the spill slope
$\beta$, moves GE up or down. At a day whose spill equals the stratum mean,
$\widehat{GE}_d = \psi_s$. This is applied **per posterior draw**, producing a
daily GE draw matrix (days × draws); all days in a draw share the same
$(\alpha, \beta, \varepsilon_s)$, so daily GE values are correlated through the
coefficients rather than carrying independent daily noise.

---

## Into SCRAPI

Each column of the daily GE draw matrix is one posterior draw. SCRAPI consumes
one column per bootstrap iteration, so GE uncertainty — wider where the PIT data
are thin — propagates into the abundance confidence intervals. (No abundance
numbers are shown here.)

---

## Modeling choices for discussion

1. **Identifiability of the GRS detection probability `p`.** It is identified
   only by the fifth cell (missed at both LGR arrays, recovered at GOJ). Under
   80–85% spill, a fish missed by GRS likely re-spills at Little Goose and is
   never seen at GOJ, shrinking that cell toward zero. Is GOJ recovery alone
   enough, or should LMJ/MCJ/IHR recoveries be added? Does the informative McCann
   prior on `phi` stabilise this, or mask weak identification?
2. **Bay-1 attraction `q_d`.** `p` currently absorbs $\psi^* \cdot q_d$ into one
   spill-dependent probability. DART's `percent_spill_spillbay1` is a candidate
   covariate for $q_d$ — worth adding?
3. **Overdispersion.** The `max_n = 500` multinomial thinning vs an explicit
   Dirichlet-multinomial / per-observation random effect that lets the data set
   its own effective sample size.
4. **Separate vs joint estimation.** GE is fit in this standalone model and the
   posterior draws are plugged into SCRAPI's $1/GE$ expansion, rather than
   estimating GE and abundance jointly (as in a full time-stratified CJS
   abundance model). Is the two-step decoupling losing coherence / understating
   uncertainty?
5. **Resolution and pooling.** Weekly strata nested in collapse strata, then
   resolved to daily via the spill regression. Is weekly-nested the right
   granularity?

---

*Analysis in R using PTAGIS interrogation data, DART daily operations records,
the `smoltEASE` package, and McCann et al. (2023) Table 8.3 coefficients. GE
model: `smoltEASE::fit_ge_model` (multistate, hierarchical, JAGS).*
