---
title: "SCRAPI2: Propagating GE and GSI uncertainty in smolt escapement estimation"
subtitle: "Validation against SCOBI::SCRAPI — MY2025 steelhead"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
vignette: >
  %\VignetteIndexEntry{SCRAPI2: GE and GSI uncertainty in smolt escapement estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, eval = TRUE)
```

---

# Background

## What SCRAPI does

SCRAPI (Steinhorst & Ackerman) estimates wild smolt escapement past Lower Granite
Dam (LGR) by expanding trap catches to full passage with a daily sampling rate:

$$\hat{N}_d = \frac{C_d}{R_d \times GE_d}$$

$C_d$ is the trap count, $R_d$ the sampling rate, and $GE_d$ the guidance
efficiency — the fraction of migrants entering the juvenile bypass. The bootstrap
resamples passage counts and trap fish, propagating uncertainty in $R_d$ and the
stock/rear composition into the CIs.

## The GE problem

Historically $GE_d$ is a **fixed** NOAA point estimate, so the CIs reflect only
trap-sampling uncertainty — not uncertainty in GE. Under post-ROD operations LGR
runs at 80–85% spill during peak migration, where GE is low and poorly
constrained, so a fixed GE badly understates the uncertainty in escapement.

## What SCRAPI2 adds

`SCRAPI2` propagates two extra sources, independently or together: **`geDraws`**
(per-day posterior GE draws from a Bayesian model) and **`gsiDraws`** (per-fish
posterior stock-assignment draws, the adult-side EASE format). Each bootstrap
iteration consumes one draw. **Guarantee (validated in §5, §7):** with degenerate
draws (every column equal, or `geDraws = NULL` + a fixed `GuidanceEfficiency`
column, and `gsiDraws = NULL`), `SCRAPI2` reproduces `SCOBI::SCRAPI` to within
bootstrap noise — it changes *what uncertainty is propagated*, not the arithmetic.

---

# Install and load

```{r install, eval=FALSE}
remotes::install_github("dsmi313/smoltEASE")
```

```{r libraries}
library(smoltEASE)
library(tidyverse)
library(lubridate)
library(readxl)
library(scales)
library(coda)        # gelman.diag

B            <- 2000                 # bootstrap iterations / draws
season_start <- ymd("2025-03-24")
season_end   <- ymd("2025-08-01")
```

---

# Data inputs

Dates are handled as `"%m/%d/%Y"` strings; `SCRAPI2`/`SCRAPI` use
`dateFormat = "%m/%d/%Y"`.

```{r files}
# Pre-formatted trap (one row per smolt) and FPC passage (Collapse strata + the
# fixed NOAA GuidanceEfficiency already merged in).
STHD.trap <- read.csv("MY2025STHD.trapData_formatted.csv")
STHD.FPC  <- read.csv("MY2025STHDsmolts_FPCdata_strata_added.csv")
STHD.trap$CollectionDate <- format(as.Date(STHD.trap$CollectionDate, "%Y-%m-%d"), "%m/%d/%Y")
STHD.FPC$SampleEndDate   <- format(as.Date(STHD.FPC$SampleEndDate,   "%m/%d/%Y"), "%m/%d/%Y")
stopifnot(all(c("Week", "Collapse", "GuidanceEfficiency", "SampleEndDate",
                "SampleCount", "SampleRate") %in% names(STHD.FPC)))

# GSI posterior draws: column 1 = MasterID, remaining columns = stock draws.
gsi_draws <- read.delim("OmyLGRJ25C_gsiDraws_final.txt", sep = "\t",
                        header = TRUE, check.names = FALSE)

# Week -> collapse-stratum map (STHD tab). Drives the GE model and the trap's
# Collapse join for legacy SCRAPI; also the parent grouping for nested weekly GE.
strata_raw  <- read_xlsx("Smolt Strata.xlsx", sheet = 1)
sthd_start  <- which(strata_raw[[1]] == "WeekNumber" & strata_raw[[2]] == "STHD_n")
sthd_strata <- strata_raw[(sthd_start + 1):nrow(strata_raw), c(1, 3)] |>
  set_names(c("Week", "Collapse")) |>
  filter(!is.na(Week)) |>
  mutate(Week = as.integer(Week), Collapse = as.integer(Collapse))
```

```{r derive}
# PIT first-detections at GRJ (bypass), GRS (spillway bay 1), GOJ (Little Goose).
pit_raw <- read_csv("MY25_STHD.csv") |>
  mutate(site     = str_extract(`Site Name`, "^[A-Z]+"),
         obs_date = as_date(parse_date_time(`Obs Time Value`,
                                            orders = c("mdy HM", "mdy HMS")))) |>
  filter(`Rear Type Code` %in% c("W", "H", "U"),
         site %in% c("GRJ", "GRS", "GOJ"),
         obs_date >= season_start, obs_date < season_end) |>
  group_by(`Tag Code`, site) |>
  slice_min(obs_date, with_ties = FALSE) |>
  ungroup() |>
  rename(tag = `Tag Code`, det_date = obs_date)

# Daily spill (% of flow): LGR drives psi/p (and the daily GE downscale); LGS phi.
lgr_spill <- read_csv("DART Spill.csv") |>
  mutate(Date = mdy(date),
         spill.per = replace_na(as.numeric(spill_kcfs) / as.numeric(outflow_kcfs) * 100, 0)) |>
  filter(Date >= season_start, Date < season_end) |>
  select(Date, spill.per)

lgs_spill <- bind_rows(read_csv("spill.all.csv"), read_csv("spill.all2.csv"),
                       read_csv("spill.all3.csv")) |>
  mutate(Date = as_date(mdy_hm(DateTime)), spill.per = HourlySpill / HourlyFlow * 100) |>
  filter(Site == "LGS", Date >= season_start, Date < season_end) |>
  group_by(Date) |>
  summarise(spill.per = mean(spill.per, na.rm = TRUE), .groups = "drop")
```

---

# The GE model — weekly, stratum-guided, daily

## What it estimates

A **multistate mark-recapture model** (after Hance et al. 2020/2024). From
GRJ/GRS/GOJ detections it jointly estimates, per stratum $s$:

- $\psi_s$ — P(fish takes the **bypass** route). **This is GE:** $GE_s=\psi_s$.
- $p_s$ — P(spillway fish **detected at GRS** bay 1).
- $\phi_s$ — P(fish **detected at GOJ** downstream).

Detected fish fall into five histories with probabilities (conditioned on
detection): $\psi(1{-}\phi)$, $\psi\phi$, $(1{-}\psi)p(1{-}\phi)$,
$(1{-}\psi)p\phi$, $(1{-}\psi)(1{-}p)\phi$. The last — missed at LGR, seen at GOJ
— makes $p$ (and hence $\psi$) identifiable. Each parameter is a hierarchical
logistic regression on spill ($\psi,p$ on LGR; $\phi$ on LGS), with
weakly-informative Student-$t$ priors ($\beta\le0$) and **informative McCann
(2023) priors on $\phi$**.

**This vignette's default is the fully-resolved model:** strata are **weeks**,
each week's $\text{logit}(\psi)$ is **nested** in (shrinks toward) its
collapse-stratum mean `mu_strat[parent]` — so sparse weeks borrow strength
instead of producing high-leverage near-zero GE — and GE is **resolved to daily**
spill. All three are opt-in flags; turning them off recovers the collapsed,
constant-GE fit.

<details><summary>Full JAGS model (nested — the default; single-level drops <code>mu_strat</code>)</summary>

```
model {
  for (s in 1:S) {                                  # s indexes weekly strata
    logit_psi[s] ~ dnorm(mu_strat[parent[s]] + beta * lgr_spill_std[s], tau_psi)  # week shrinks to its collapse-stratum mean
    logit_p[s]   ~ dnorm(alpha_p   + beta_p   * lgr_spill_std[s], tau_p)
    logit_phi[s] ~ dnorm(alpha_phi + beta_phi * lgs_spill_std[s], tau_phi)
    psi[s] <- ilogit(logit_psi[s])      # psi == GE
    p[s]   <- ilogit(logit_p[s]); phi[s] <- ilogit(logit_phi[s])
    pi_raw[s,1] <- psi[s]*(1-phi[s]);           pi_raw[s,2] <- psi[s]*phi[s]
    pi_raw[s,3] <- (1-psi[s])*p[s]*(1-phi[s]);  pi_raw[s,4] <- (1-psi[s])*p[s]*phi[s]
    pi_raw[s,5] <- (1-psi[s])*(1-p[s])*phi[s]
    pi_obs[s,] <- pi_raw[s,] / sum(pi_raw[s,])
  }
  for (j in 1:S_lik) n[lik_idx[j],1:5] ~ dmulti(pi_obs[lik_idx[j],1:5], N_seen[lik_idx[j]])

  for (g in 1:n_strat) mu_strat[g] ~ dnorm(alpha, tau_strat)   # collapse-stratum means shrink to the global mean
  alpha ~ dt(0,pow(2,-2),7); beta ~ dt(0,pow(1,-2),7) T(,0)
  sigma_psi ~ dunif(0.05,3); sigma_strat ~ dunif(0.05,3)
  alpha_p ~ dt(0,pow(2,-2),7); beta_p ~ dt(0,pow(1,-2),7);   sigma_p   ~ dunif(0.05,3)
  alpha_phi ~ dnorm(-2.60,pow(0.30,-2)); beta_phi ~ dnorm(-1.78,pow(0.25,-2)) T(,0)
  sigma_phi ~ dunif(0.05,1)
  # single-level (no parent_strata): logit_psi[s] ~ dnorm(alpha + beta*lgr_spill_std[s], tau_psi); drop mu_strat/sigma_strat
}
```

</details>

## Fit (weekly, nested, daily — the default)

```{r ge-fit}
sthd_weekly <- transform(sthd_strata, Collapse = Week)   # one stratum per week

ge_data <- prep_ge_data(
  dat_up = pit_raw, strat_assign = sthd_weekly,
  spill_data = lgr_spill, lgs_spill_data = lgs_spill,
  species = "sthd", goj_site = "GOJ",
  parent_strata = sthd_strata                            # weeks nested in collapse strata
)

ge_fit <- fit_ge_model(ge_data, max_n = 500, seed = 42)  # nesting auto-detected

# Daily-resolved GE draws (one column per bootstrap iteration)
ge_draws <- generate_ge_draws(
  ge_fit, strat_assign = sthd_weekly,
  pass_dates  = as.Date(STHD.FPC$SampleEndDate, "%m/%d/%Y"),
  B           = B,
  daily_spill = lgr_spill
)

# Convergence: psi (= GE), parent means, and the regression scalars; want < 1.05
mon <- grep("^psi\\[|^mu_strat\\[|^alpha$|^beta$|^sigma_psi$|^sigma_strat$",
            coda::varnames(ge_fit$samples))
gelman.diag(ge_fit$samples[, mon], autoburnin = FALSE, multivariate = FALSE)

stopifnot(nrow(ge_draws) == nrow(STHD.FPC),
          all(unlist(ge_draws[, -1]) >= 0), all(unlist(ge_draws[, -1]) <= 1))
```

## Stratum-guided weekly GE

One picture of the whole GE model: the **weekly** posterior GE (point + 95%
ribbon) tracks daily spill, while shrinking toward its **collapse-stratum** mean
(the parent it is guided by). Both sit well below the **NOAA fixed** GE once
spill ramps up.

```{r ge-diagnostic, fig.width=9, fig.height=4.5}
ge_mat <- as.matrix(ge_draws[, -1])
ge_day <- tibble(
  date     = as.Date(ge_draws$SampleEndDate, "%m/%d/%Y"),
  Collapse = STHD.FPC$Collapse,
  weekly   = rowMeans(ge_mat),
  lo       = apply(ge_mat, 1, quantile, 0.025),
  hi       = apply(ge_mat, 1, quantile, 0.975),
  NOAA     = STHD.FPC$GuidanceEfficiency
) |>
  group_by(Collapse) |>
  mutate(stratum = mean(weekly)) |>     # parent collapse-stratum GE
  ungroup()

ggplot(ge_day, aes(date)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "#d73027", alpha = 0.15) +
  geom_line(aes(y = weekly,  colour = "Weekly model GE (daily)"), linewidth = 0.9) +
  geom_line(aes(y = stratum, colour = "Collapse-stratum GE (parent)"), linewidth = 0.8) +
  geom_line(aes(y = NOAA,    colour = "NOAA fixed GE"), linetype = "dashed") +
  scale_colour_manual(values = c("Weekly model GE (daily)"     = "#d73027",
                                 "Collapse-stratum GE (parent)" = "#2166ac",
                                 "NOAA fixed GE"                = "grey30")) +
  scale_y_continuous("Guidance efficiency", labels = percent_format(accuracy = 1)) +
  labs(x = NULL, colour = NULL,
       title = "Stratum-guided weekly GE vs NOAA fixed — MY2025") +
  theme_minimal(base_size = 13) + theme(legend.position = "top")
```

---

# Three-way comparison

Same data, three GE treatments. **A vs B** validates that `SCRAPI2` did not change
the SCRAPI arithmetic; **C** is the weekly+nested+daily default and should have
**wider** CIs (GE uncertainty propagated).

| Workflow | GE source | Tool |
|---|---|---|
| **A — Legacy** | NOAA `GuidanceEfficiency` (in FPC) | `SCOBI::SCRAPI` |
| **B — SCRAPI2 fixed GE** | Same NOAA column, no uncertainty | `SCRAPI2` (`geDraws = NULL`) |
| **C — SCRAPI2 model GE** | Weekly + nested + daily draws | `SCRAPI2` (`geDraws = ge_draws`) |

```{r scrapi2-args, include=FALSE}
# shared SCRAPI2 column mapping / settings
s2 <- function(...) SCRAPI2(
  passageData = STHD.FPC, Primary = "GenStock",
  strat = "Week", dat = "SampleEndDate", tally = "SampleCount",
  samrate = "SampleRate", guidance = "GuidanceEfficiency", collaps = "Collapse",
  RTYPE = "W", alph = 0.05, dateFormat = "%m/%d/%Y", B = B, ...)
```

## Path A — Legacy `SCOBI::SCRAPI`

`SCOBI::SCRAPI` needs `Collapse` in both frames. The FPC has it; map it onto the
trap from `sthd_strata`.

```{r path-a}
add_collapse <- function(df, date_col, strata_tbl) {
  d <- as.Date(df[[date_col]], "%m/%d/%Y")
  df$Week <- isoweek(d); df <- left_join(df, strata_tbl, by = "Week")
  df[[date_col]] <- format(d, "%m/%d/%Y"); df
}
STHD.trap_legacy <- add_collapse(STHD.trap, "CollectionDate", sthd_strata)
stopifnot(!anyNA(STHD.trap_legacy$Collapse), !anyNA(STHD.FPC$Collapse))

SCOBI::SCRAPI(smoltData = STHD.trap_legacy, passageData = STHD.FPC,
              Primary = "GenStock", dateFormat = "%m/%d/%Y", Run = "legacy_", B = B)

legacy_ci <- read.csv("legacy_CIs.csv", skip = 1, header = TRUE,
                      row.names = 1, check.names = FALSE) %>%
  filter(!grepl("SampleSize", rownames(.)))
```

## Paths B and C — `SCRAPI2`

```{r paths-bc}
result_fixed <- s2(smoltData = STHD.trap, Run = "fixed_")                 # fixed NOAA GE
result_model <- s2(smoltData = STHD.trap, geDraws = ge_draws, Run = "model_")  # weekly+nested+daily
```

---

# Results

```{r results-table}
ci_df <- function(mat, label)
  as.data.frame(mat) %>% rownames_to_column("Stock") %>% mutate(Workflow = label)

results_all <- bind_rows(
  legacy_ci %>% rownames_to_column("Stock") %>% mutate(Workflow = "A: SCOBI legacy"),
  ci_df(result_fixed$CI, "B: SCRAPI2 fixed GE"),
  ci_df(result_model$CI, "C: SCRAPI2 model GE")
)

results_all %>%
  filter(Stock == "WildSmolts") %>%
  mutate(`CI half-width (%)` = round((UCI - LCI) / (2 * Estimate) * 100, 1)) %>%
  select(Workflow, Estimate, LCI, UCI, `CI half-width (%)`) %>%
  knitr::kable(caption = "Wild steelhead smolt estimates — three workflows",
               format.args = list(big.mark = ","))
```

```{r forest-plot, fig.width=10, fig.height=5}
pal <- c("A: SCOBI legacy" = "#555555", "B: SCRAPI2 fixed GE" = "#2166ac",
         "C: SCRAPI2 model GE" = "#d73027")
results_all %>%
  filter(Stock != "WildSmolts") %>%
  ggplot(aes(Estimate, factor(Stock), xmin = LCI, xmax = UCI, colour = Workflow)) +
  geom_pointrange(position = position_dodge(width = 0.5), size = 0.6) +
  scale_colour_manual(values = pal) + scale_x_continuous(labels = comma) +
  labs(title = "Wild steelhead smolts by genetic stock — MY2025",
       x = "Estimated abundance", y = NULL, colour = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", axis.text.y = element_text(hjust = 1))
```

```{r ci-width-plot, fig.width=6, fig.height=4}
results_all %>%
  filter(Stock == "WildSmolts") %>%
  mutate(half_width = (UCI - LCI) / 2, Workflow = fct_inorder(Workflow)) %>%
  ggplot(aes(Workflow, half_width, fill = Workflow)) +
  geom_col(width = 0.5, show.legend = FALSE) + scale_fill_manual(values = pal) +
  labs(title = "CI half-width — WildSmolts", x = NULL, y = "Half-width (fish)") +
  theme_minimal(base_size = 13)
```

A ≈ B (SCRAPI ≡ SCRAPI2 with fixed GE); C is wider — the extra width is the GE
uncertainty the weekly model propagates.

---

# GSI uncertainty: fixed vs drawn

GE uncertainty widens the **total** CI; GSI uncertainty widens the **per-stock**
CIs without changing the total (rear composition and passage are untouched). We
hold GE fixed, restrict to genotyped fish, and compare fixed (`gsiDraws = NULL`)
vs drawn (`gsiDraws = gsi_draws`).

```{r gsi-runs}
gsi_ids       <- gsi_draws[[1]]
STHD.trap.gsi <- STHD.trap[STHD.trap$MasterID %in% gsi_ids, ]
cat(sprintf("Trap fish: %d total, %d genotyped, %d excluded\n",
            nrow(STHD.trap), nrow(STHD.trap.gsi), nrow(STHD.trap) - nrow(STHD.trap.gsi)))

result_gsi_fixed <- s2(smoltData = STHD.trap.gsi, Run = "gsifixed_")
result_gsi       <- s2(smoltData = STHD.trap.gsi, gsiDraws = gsi_draws,
                       fishID = "MasterID", Run = "gsidrawn_")
```

```{r gsi-total-check}
cat(sprintf("WildSmolts total  fixed GSI: %s   drawn GSI: %s\n",
            format(result_gsi_fixed$CI["WildSmolts", "Estimate"], big.mark = ","),
            format(result_gsi$CI["WildSmolts",       "Estimate"], big.mark = ",")))
```

```{r gsi-compare-table}
bind_rows(ci_df(result_gsi_fixed$CI, "Fixed"), ci_df(result_gsi$CI, "Drawn")) %>%
  filter(Stock != "WildSmolts") %>%
  transmute(Stock, Workflow, Estimate = round(Estimate),
            HalfWidth = round((UCI - LCI) / 2)) %>%
  pivot_wider(names_from = Workflow, values_from = c(Estimate, HalfWidth)) %>%
  mutate(Width_ratio = round(HalfWidth_Drawn / HalfWidth_Fixed, 2)) %>%
  knitr::kable(caption = paste("Per-stock smolts: fixed vs drawn GSI (genotyped",
               "fish, GE fixed). Width_ratio > 1 = genetic uncertainty widens the CI."),
               format.args = list(big.mark = ","))
```

```{r gsi-forest, fig.width=10, fig.height=5}
bind_rows(ci_df(result_gsi_fixed$CI, "Fixed GSI"), ci_df(result_gsi$CI, "Drawn GSI")) %>%
  filter(Stock != "WildSmolts") %>%
  ggplot(aes(Estimate, factor(Stock), xmin = LCI, xmax = UCI, colour = Workflow)) +
  geom_pointrange(position = position_dodge(width = 0.5), size = 0.6) +
  scale_colour_manual(values = c("Fixed GSI" = "#2166ac", "Drawn GSI" = "#d73027")) +
  scale_x_continuous(labels = comma) +
  labs(title = "Per-stock smolts: fixed vs drawn GSI (GE fixed) — MY2025",
       x = "Estimated abundance", y = NULL, colour = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", axis.text.y = element_text(hjust = 1))
```

The total matches; per-stock CIs widen under drawn GSI (`Width_ratio > 1`). GE and
GSI draws can be supplied together for full propagation. These subset totals need
not match the full-trap GE comparison above.

---

# Control checks (pathway neutrality)

Both draw pathways must be numerically neutral when the draws carry no
uncertainty — the strict identity test.

```{r control-exact-ge, results='hold'}
# Exact-GE: every draw equals the fixed NOAA GE -> must match the fixed-GE run.
ge0 <- pmin(pmax(as.numeric(STHD.FPC$GuidanceEfficiency), 1e-6), 1 - 1e-6)
ge_exact <- data.frame(SampleEndDate = STHD.FPC$SampleEndDate,
                       matrix(rep(ge0, B), nrow = length(ge0), ncol = B),
                       check.names = FALSE)
result_exact <- s2(smoltData = STHD.trap, geDraws = ge_exact, Run = "exact_")
d_ge <- abs(result_fixed$CI["WildSmolts","Estimate"] - result_exact$CI["WildSmolts","Estimate"]) /
        result_fixed$CI["WildSmolts","Estimate"]
cat(sprintf("Exact-GE control  WildSmolts diff: %.2f%%\n", 100 * d_ge))
stopifnot(d_ge < 0.01)
```

```{r control-degenerate-gsi, results='hold'}
# Degenerate-GSI: observed GenStock repeated as draws -> must match gsiDraws = NULL.
gsi_degen <- sim_gsi_draws(STHD.trap.gsi, Primary = "GenStock", fishID = "MasterID",
                           B = B, method = "fixed")
result_gsi_degen <- s2(smoltData = STHD.trap.gsi, gsiDraws = gsi_degen,
                       fishID = "MasterID", Run = "gsidegen_")
d_gsi <- abs(result_gsi_fixed$CI["WildSmolts","Estimate"] - result_gsi_degen$CI["WildSmolts","Estimate"]) /
         result_gsi_fixed$CI["WildSmolts","Estimate"]
cat(sprintf("Degenerate-GSI control  WildSmolts diff: %.2f%%\n", 100 * d_gsi))
stopifnot(d_gsi < 0.01)
```

---

# Reviewer checklist

| Check | Expected result |
|---|---|
| `gelman.diag()` R-hat (psi, mu_strat, alpha, beta, sigma_psi/strat) | < 1.05 for all |
| GE draws range | All in \[0, 1\] |
| Stratum-guided weekly GE plot | Weekly GE shrinks toward its collapse-stratum line; data-rich weeks barely move, sparse weeks pulled toward the parent; both below NOAA at high spill |
| **Path A vs Path B** | **WildSmolts within ~5% (bootstrap noise) — SCRAPI ≡ SCRAPI2** |
| **Exact-GE control** | **< 1% WildSmolts difference — geDraws pathway neutral** |
| **Degenerate-GSI control** | **< 1% WildSmolts difference — gsiDraws pathway neutral** |
| Path C CI vs Path B | Wider or equal (GE uncertainty) |
| Fixed vs drawn GSI | Total unchanged; per-stock CIs wider (`Width_ratio` > 1) |

---
