---
title: "SCRAPI2: Propagating GE and GSI uncertainty in smolt escapement estimation"
subtitle: "Validation against SCOBI::SCRAPI — MY2025 steelhead"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
vignette: >
  %\VignetteIndexEntry{SCRAPI2: GE and GSI uncertainty in smolt escapement estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, eval = TRUE)
```

# Background

## What SCRAPI does

SCRAPI (Steinhorst & Ackerman) estimates wild smolt escapement past Lower Granite
Dam (LGR) by expanding trap catches to full passage with a daily sampling rate:

$$\hat{N}_d = \frac{C_d}{R_d \times GE_d}$$

$C_d$ is the trap count, $R_d$ the sampling rate, and $GE_d$ the guidance
efficiency — the fraction of migrants entering the juvenile bypass. The bootstrap
resamples passage counts and trap fish, propagating uncertainty in $R_d$ and the
stock/rear composition into the CIs.

## The GE problem

Historically $GE_d$ is a **fixed** NOAA point estimate, so the CIs reflect only
trap-sampling uncertainty — not uncertainty in GE. Under post-ROD operations LGR
runs at 80–85% spill during peak migration, where GE is low and poorly
constrained, so a fixed GE badly understates the uncertainty in escapement.

## What SCRAPI2 adds

`SCRAPI2` propagates two additional uncertainty sources, independently or
together: **`geDraws`** (per-day posterior GE draws from a Bayesian model) and
**`gsiDraws`** (per-fish posterior stock-assignment draws, the adult-side EASE
format). Each bootstrap iteration consumes one draw. The arithmetic is unchanged
from `SCOBI::SCRAPI` — what changes is which uncertainty sources are propagated
into the CIs. This guarantee is verified by two control checks: an exact-GE
control (§8) and a degenerate-GSI control (§8), both of which confirm < 1%
deviation from the fixed-GE run before results are reported.

---

# Install and load

```{r install, eval=FALSE}
remotes::install_github("dsmi313/smoltEASE")
```

```{r libraries}
library(smoltEASE)
library(tidyverse)
library(lubridate)
library(readxl)
library(scales)
library(coda)

B            <- 2000
season_start <- ymd("2025-03-24")
season_end   <- ymd("2025-08-01")

# Shared SCRAPI2 column mapping used throughout; defined here so it is visible
# before first use.
s2 <- function(...) SCRAPI2(
  passageData = STHD.FPC, Primary = "GenStock",
  strat = "Week", dat = "SampleEndDate", tally = "SampleCount",
  samrate = "SampleRate", guidance = "GuidanceEfficiency", collaps = "Collapse",
  RTYPE = "W", alph = 0.05, dateFormat = "%m/%d/%Y", B = B, ...)
```

---

# Data inputs

Dates are handled as `"%m/%d/%Y"` strings throughout; `SCRAPI2`/`SCRAPI` use
`dateFormat = "%m/%d/%Y"`.

```{r files}
# Trap (one row per smolt) and FPC passage data. The FPC already has Collapse
# strata and the fixed NOAA GuidanceEfficiency merged in.
STHD.trap <- read.csv("MY2025STHD.trapData_formatted.csv")
STHD.FPC  <- read.csv("MY2025STHDsmolts_FPCdata_strata_added.csv")
STHD.trap$CollectionDate <- format(as.Date(STHD.trap$CollectionDate, "%Y-%m-%d"), "%m/%d/%Y")
STHD.FPC$SampleEndDate   <- format(as.Date(STHD.FPC$SampleEndDate,   "%m/%d/%Y"), "%m/%d/%Y")
stopifnot(all(c("Week", "Collapse", "GuidanceEfficiency", "SampleEndDate",
                "SampleCount", "SampleRate") %in% names(STHD.FPC)))

# GSI posterior draws: column 1 = MasterID, remaining columns = stock draws.
gsi_draws <- read.delim("OmyLGRJ25C_gsiDraws_final.txt", sep = "\t",
                        header = TRUE, check.names = FALSE)

# Week -> collapse-stratum map. The strata workbook has a header block above the
# STHD data; locate the WeekNumber row dynamically to avoid hard-coding a skip value.
strata_raw  <- read_xlsx("Smolt Strata.xlsx", sheet = 1)
sthd_start  <- which(strata_raw[[1]] == "WeekNumber" & strata_raw[[2]] == "STHD_n")
sthd_strata <- strata_raw[(sthd_start + 1):nrow(strata_raw), c(1, 3)] |>
  set_names(c("Week", "Collapse")) |>
  filter(!is.na(Week)) |>
  mutate(Week = as.integer(Week), Collapse = as.integer(Collapse))
```

```{r derive}
# PIT first-detections at GRJ (bypass), GRS (spillway bay 1), GOJ (Little Goose).
pit_raw <- read_csv("MY25_STHD.csv") |>
  mutate(site     = str_extract(`Site Name`, "^[A-Z]+"),
         obs_date = as_date(parse_date_time(`Obs Time Value`,
                                            orders = c("mdy HM", "mdy HMS")))) |>
  filter(`Rear Type Code` %in% c("W", "H", "U"),
         site %in% c("GRJ", "GRS", "GOJ"),
         obs_date >= season_start, obs_date < season_end) |>
  group_by(`Tag Code`, site) |>
  slice_min(obs_date, with_ties = FALSE) |>
  ungroup() |>
  rename(tag = `Tag Code`, det_date = obs_date)

# LGR daily spill (% of flow): drives psi, p, and the daily GE downscale.
lgr_spill <- read_csv("DART Spill.csv") |>
  mutate(Date      = mdy(date),
         spill.per = replace_na(as.numeric(spill_kcfs) / as.numeric(outflow_kcfs) * 100, 0)) |>
  filter(Date >= season_start, Date < season_end) |>
  select(Date, spill.per)

# LGS daily spill: drives phi. Sourced from three sequential DART query exports
# (DART limits row counts per download).
lgs_spill <- bind_rows(read_csv("spill.all.csv"),
                       read_csv("spill.all2.csv"),
                       read_csv("spill.all3.csv")) |>
  mutate(Date      = as_date(mdy_hm(DateTime)),
         spill.per = HourlySpill / HourlyFlow * 100) |>
  filter(Site == "LGS", Date >= season_start, Date < season_end) |>
  group_by(Date) |>
  summarise(spill.per = mean(spill.per, na.rm = TRUE), .groups = "drop")
```

---

# Control checks (pathway neutrality) {#controls}

Both draw pathways must be numerically neutral when the draws carry no uncertainty.
These checks are run before results so that if either fails, results are not
reported.

```{r control-exact-ge, results='hold'}
# Exact-GE: every draw equals the fixed NOAA GE -> must match the fixed-GE run.
ge0 <- pmin(pmax(as.numeric(STHD.FPC$GuidanceEfficiency), 1e-6), 1 - 1e-6)
ge_exact <- data.frame(SampleEndDate = STHD.FPC$SampleEndDate,
                       matrix(rep(ge0, B), nrow = length(ge0), ncol = B),
                       check.names = FALSE)
result_fixed_ctrl <- s2(smoltData = STHD.trap, geDraws = NULL,  Run = "fixed_ctrl_")
result_exact      <- s2(smoltData = STHD.trap, geDraws = ge_exact, Run = "exact_")
d_ge <- abs(result_fixed_ctrl$CI["WildSmolts", "Estimate"] -
              result_exact$CI["WildSmolts",    "Estimate"]) /
        result_fixed_ctrl$CI["WildSmolts", "Estimate"]
cat(sprintf("Exact-GE control  WildSmolts diff: %.2f%%\n", 100 * d_ge))
stopifnot(d_ge < 0.01)
```

```{r control-degenerate-gsi, results='hold'}
# Degenerate-GSI: observed GenStock repeated as draws -> must match gsiDraws = NULL.
gsi_ids       <- gsi_draws[[1]]
STHD.trap.gsi <- STHD.trap[STHD.trap$MasterID %in% gsi_ids, ]
cat(sprintf("Trap fish: %d total, %d genotyped, %d excluded\n",
            nrow(STHD.trap), nrow(STHD.trap.gsi), nrow(STHD.trap) - nrow(STHD.trap.gsi)))

result_gsi_fixed <- s2(smoltData = STHD.trap.gsi, Run = "gsifixed_")
gsi_degen        <- sim_gsi_draws(STHD.trap.gsi, Primary = "GenStock",
                                  fishID = "MasterID", B = B, method = "fixed")
result_gsi_degen <- s2(smoltData = STHD.trap.gsi, gsiDraws = gsi_degen,
                       fishID = "MasterID", Run = "gsidegen_")
d_gsi <- abs(result_gsi_fixed$CI["WildSmolts", "Estimate"] -
               result_gsi_degen$CI["WildSmolts", "Estimate"]) /
         result_gsi_fixed$CI["WildSmolts", "Estimate"]
cat(sprintf("Degenerate-GSI control  WildSmolts diff: %.2f%%\n", 100 * d_gsi))
stopifnot(d_gsi < 0.01)
```

---

# The GE model — weekly, stratum-guided, daily

## What it estimates

A **multistate mark-recapture model** (after Hance et al. 2020/2024). From
GRJ/GRS/GOJ detections it jointly estimates, per stratum $s$:

- $\psi_s$ — P(fish takes the **bypass** route). **This is GE:** $GE_s = \psi_s$.
- $p_s$ — P(spillway fish **detected at GRS** bay 1).
- $\phi_s$ — P(fish **detected at GOJ** downstream).

Detected fish fall into five histories (conditioned on detection):

| History | Route | Probability |
|---|---|---|
| GRJ, not GOJ | Bypass, missed downstream | $\psi(1-\phi)$ |
| GRJ and GOJ | Bypass, seen downstream | $\psi\phi$ |
| GRS, not GOJ | Spillway, detected at bay 1, missed downstream | $(1-\psi)\,p\,(1-\phi)$ |
| GRS and GOJ | Spillway, detected at bay 1, seen downstream | $(1-\psi)\,p\,\phi$ |
| No LGR, GOJ | Spillway, **missed** at bay 1, seen downstream | $(1-\psi)(1-p)\phi$ |

The fifth history — missed at both LGR arrays but recovered at GOJ — is what makes $p$ (and hence $\psi$) identifiable.
Each parameter is a hierarchical logistic regression on spill ($\psi, p$ on LGR;
$\phi$ on LGS), with weakly-informative Student-$t$ priors ($\beta \le 0$) and
**informative McCann (2023) priors on $\phi$**.

The default fit uses **weekly strata** with each week's $\text{logit}(\psi)$
**nested** in its collapse-stratum mean `mu_strat[parent]`, so sparse weeks borrow
strength rather than producing high-leverage near-zero GE estimates. GE is then
**resolved to daily** values via the spill regression. The three options —
weekly strata, nesting, daily resolution — are each opt-in flags; disabling all
three recovers the collapsed, constant-GE fit.

<details><summary>Full JAGS model (nested — the default; single-level drops <code>mu_strat</code>)</summary>

```
model {
  for (s in 1:S) {
    logit_psi[s] ~ dnorm(mu_strat[parent[s]] + beta * lgr_spill_std[s], tau_psi)
    logit_p[s]   ~ dnorm(alpha_p   + beta_p   * lgr_spill_std[s], tau_p)
    logit_phi[s] ~ dnorm(alpha_phi + beta_phi * lgs_spill_std[s], tau_phi)
    psi[s] <- ilogit(logit_psi[s])
    p[s]   <- ilogit(logit_p[s]); phi[s] <- ilogit(logit_phi[s])
    pi_raw[s,1] <- psi[s]*(1-phi[s]);           pi_raw[s,2] <- psi[s]*phi[s]
    pi_raw[s,3] <- (1-psi[s])*p[s]*(1-phi[s]);  pi_raw[s,4] <- (1-psi[s])*p[s]*phi[s]
    pi_raw[s,5] <- (1-psi[s])*(1-p[s])*phi[s]
    pi_obs[s,] <- pi_raw[s,] / sum(pi_raw[s,])
  }
  for (j in 1:S_lik) n[lik_idx[j],1:5] ~ dmulti(pi_obs[lik_idx[j],1:5], N_seen[lik_idx[j]])

  for (g in 1:n_strat) mu_strat[g] ~ dnorm(alpha, tau_strat)
  alpha ~ dt(0,pow(2,-2),7); beta ~ dt(0,pow(1,-2),7) T(,0)
  sigma_psi ~ dunif(0.05,3); sigma_strat ~ dunif(0.05,3)
  alpha_p ~ dt(0,pow(2,-2),7); beta_p ~ dt(0,pow(1,-2),7); sigma_p ~ dunif(0.05,3)
  alpha_phi ~ dnorm(-2.60,pow(0.30,-2)); beta_phi ~ dnorm(-1.78,pow(0.25,-2)) T(,0)
  sigma_phi ~ dunif(0.05,1)
  # single-level: logit_psi[s] ~ dnorm(alpha + beta*lgr_spill_std[s], tau_psi); drop mu_strat/sigma_strat
}
```

</details>

## Fit

```{r ge-fit}
sthd_weekly <- transform(sthd_strata, Collapse = Week)

ge_data <- prep_ge_data(
  dat_up         = pit_raw,
  strat_assign   = sthd_weekly,
  spill_data     = lgr_spill,
  lgs_spill_data = lgs_spill,
  species        = "sthd",
  goj_site       = "GOJ",
  parent_strata  = sthd_strata
)

ge_fit <- fit_ge_model(ge_data, max_n = 500, seed = 42)

ge_draws <- generate_ge_draws(
  ge_fit,
  strat_assign = sthd_weekly,
  pass_dates   = as.Date(STHD.FPC$SampleEndDate, "%m/%d/%Y"),
  B            = B,
  daily_spill  = lgr_spill
)

mon <- grep("^psi\\[|^mu_strat\\[|^alpha$|^beta$|^sigma_psi$|^sigma_strat$",
            coda::varnames(ge_fit$samples))
gelman.diag(ge_fit$samples[, mon], autoburnin = FALSE, multivariate = FALSE)

stopifnot(
  nrow(ge_draws) == nrow(STHD.FPC),
  all(unlist(ge_draws[, -1]) >= 0),
  all(unlist(ge_draws[, -1]) <= 1),
  all(is.finite(unlist(ge_draws[, -1])))
)
```

## Stratum-guided weekly GE

The weekly posterior GE (point + 95% ribbon) tracks daily spill while shrinking
toward its collapse-stratum mean. Both sit well below the NOAA fixed GE once spill
ramps up.

```{r ge-diagnostic, fig.width=9, fig.height=4.5}
ge_mat <- as.matrix(ge_draws[, -1])
ge_day <- tibble(
  date     = as.Date(ge_draws$SampleEndDate, "%m/%d/%Y"),
  Collapse = STHD.FPC$Collapse,
  weekly   = rowMeans(ge_mat),
  lo       = apply(ge_mat, 1, quantile, 0.025),
  hi       = apply(ge_mat, 1, quantile, 0.975),
  NOAA     = STHD.FPC$GuidanceEfficiency
) |>
  group_by(Collapse) |>
  mutate(stratum = mean(weekly)) |>
  ungroup()

ggplot(ge_day, aes(date)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "#d73027", alpha = 0.15) +
  geom_line(aes(y = weekly,  colour = "Weekly model GE (daily)"),      linewidth = 0.9) +
  geom_line(aes(y = stratum, colour = "Collapse-stratum GE (parent)"), linewidth = 0.8) +
  geom_line(aes(y = NOAA,    colour = "NOAA fixed GE"), linetype = "dashed") +
  scale_colour_manual(values = c("Weekly model GE (daily)"      = "#d73027",
                                 "Collapse-stratum GE (parent)" = "#2166ac",
                                 "NOAA fixed GE"                = "grey30")) +
  scale_y_continuous("Guidance efficiency", labels = percent_format(accuracy = 1)) +
  labs(x = NULL, colour = NULL,
       title = "Stratum-guided weekly GE vs NOAA fixed — MY2025") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

---

# Three-way comparison

Same data, three GE treatments. **A vs B** confirms `SCRAPI2` did not alter the
SCRAPI arithmetic (see also the exact-GE control in §4). **C** is the
weekly + nested + daily default and should have wider CIs as GE uncertainty is
propagated.

| Workflow | GE source | Tool |
|---|---|---|
| **A — Legacy** | NOAA `GuidanceEfficiency` (in FPC) | `SCOBI::SCRAPI` |
| **B — SCRAPI2 fixed GE** | Same NOAA column, no uncertainty | `SCRAPI2` (`geDraws = NULL`) |
| **C — SCRAPI2 model GE** | Weekly + nested + daily draws | `SCRAPI2` (`geDraws = ge_draws`) |

## Path A — Legacy `SCOBI::SCRAPI`

`SCOBI::SCRAPI` requires `Collapse` in both the trap and FPC frames. The FPC
already has it; the trap gets it via a week-number join.

```{r path-a}
STHD.trap_legacy <- STHD.trap |>
  mutate(Week = isoweek(as.Date(CollectionDate, "%m/%d/%Y"))) |>
  left_join(sthd_strata, by = "Week") |>
  mutate(CollectionDate = format(as.Date(CollectionDate, "%m/%d/%Y"), "%m/%d/%Y"))

stopifnot(!anyNA(STHD.trap_legacy$Collapse), !anyNA(STHD.FPC$Collapse))

SCOBI::SCRAPI(smoltData = STHD.trap_legacy, passageData = STHD.FPC,
              Primary = "GenStock", dateFormat = "%m/%d/%Y", Run = "legacy_", B = B)

legacy_ci <- read.csv("legacy_CIs.csv", skip = 1, header = TRUE,
                      row.names = 1, check.names = FALSE, fill = TRUE)
legacy_ci <- legacy_ci[!grepl("SampleSize", rownames(legacy_ci)), ]
```

## Paths B and C — `SCRAPI2`

```{r paths-bc}
result_fixed <- s2(smoltData = STHD.trap, Run = "fixed_")
result_model <- s2(smoltData = STHD.trap, geDraws = ge_draws, Run = "model_")
```

---

# Results

```{r results-table}
ci_df <- function(mat, label)
  as.data.frame(mat) |> rownames_to_column("Stock") |> mutate(Workflow = label)

results_all <- bind_rows(
  legacy_ci |> rownames_to_column("Stock") |> mutate(Workflow = "A: SCOBI legacy"),
  ci_df(result_fixed$CI, "B: SCRAPI2 fixed GE"),
  ci_df(result_model$CI, "C: SCRAPI2 model GE")
)

results_all |>
  filter(Stock == "WildSmolts") |>
  mutate(`CI half-width (%)` = round((UCI - LCI) / (2 * Estimate) * 100, 1)) |>
  select(Workflow, Estimate, LCI, UCI, `CI half-width (%)`) |>
  knitr::kable(caption = "Wild steelhead smolt estimates — three workflows",
               format.args = list(big.mark = ","), digits = 0)
```

A ≈ B (SCRAPI ≡ SCRAPI2 with fixed GE); C is wider — the extra width is GE
uncertainty propagated from the weekly model. The upper tail of path C's CI is
notably longer than the lower tail. This asymmetry is expected: the 1/GE expansion
is convex, so low and uncertain GE values produce right-skewed abundance draws.
Weeks with the most uncertainty (low spill-adjacent GE, sparse PIT data) drive the
upper tail disproportionately.

```{r forest-plot, fig.width=10, fig.height=5}
pal <- c("A: SCOBI legacy"     = "#555555",
         "B: SCRAPI2 fixed GE" = "#2166ac",
         "C: SCRAPI2 model GE" = "#d73027")

results_all |>
  filter(Stock != "WildSmolts") |>
  ggplot(aes(Estimate, factor(Stock), xmin = LCI, xmax = UCI, colour = Workflow)) +
  geom_pointrange(position = position_dodge(width = 0.7), size = 0.5,
                  linewidth = 0.7) +
  scale_colour_manual(values = pal) +
  scale_x_continuous(labels = comma) +
  labs(title = "Wild steelhead smolts by genetic stock — MY2025",
       x = "Estimated abundance", y = NULL, colour = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", axis.text.y = element_text(hjust = 1))
```

```{r ci-width-plot, fig.width=6, fig.height=4}
results_all |>
  filter(Stock == "WildSmolts") |>
  mutate(half_width = (UCI - LCI) / 2, Workflow = fct_inorder(Workflow)) |>
  ggplot(aes(Workflow, half_width, fill = Workflow)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  scale_fill_manual(values = pal) +
  labs(title = "CI half-width — WildSmolts", x = NULL, y = "Half-width (fish)") +
  theme_minimal(base_size = 13)
```

---

# GSI uncertainty: fixed vs drawn

GE uncertainty widens the **total** CI; GSI uncertainty widens the **per-stock**
CIs without changing the total (rear composition and passage counts are untouched).
Holding GE fixed, we compare fixed (`gsiDraws = NULL`) vs drawn (`gsiDraws =
gsi_draws`) on the genotyped subset. Note that `STHD.trap.gsi` and
`result_gsi_fixed` were already computed in §4.

```{r gsi-runs}
result_gsi <- s2(smoltData = STHD.trap.gsi, gsiDraws = gsi_draws,
                 fishID = "MasterID", Run = "gsidrawn_")
```

```{r gsi-total-check}
cat(sprintf("WildSmolts total  fixed GSI: %s   drawn GSI: %s\n",
            format(result_gsi_fixed$CI["WildSmolts", "Estimate"], big.mark = ","),
            format(result_gsi$CI["WildSmolts",       "Estimate"], big.mark = ",")))
```

```{r gsi-compare-table}
bind_rows(ci_df(result_gsi_fixed$CI, "Fixed"),
          ci_df(result_gsi$CI,       "Drawn")) |>
  filter(Stock != "WildSmolts") |>
  transmute(Stock, Workflow,
            Estimate  = round(Estimate),
            HalfWidth = round((UCI - LCI) / 2)) |>
  pivot_wider(names_from = Workflow, values_from = c(Estimate, HalfWidth)) |>
  mutate(Width_ratio = round(HalfWidth_Drawn / HalfWidth_Fixed, 2)) |>
  knitr::kable(
    caption = paste("Per-stock smolts: fixed vs drawn GSI (genotyped fish, GE fixed).",
                    "Width_ratio > 1 = genetic uncertainty widens the CI.",
                    "LOCLWR and LOSALM have the largest ratios (>2), reflecting sparse",
                    "trap representation and high assignment uncertainty for those stocks."),
    format.args = list(big.mark = ","), digits = 0)
```

```{r gsi-forest, fig.width=10, fig.height=5}
bind_rows(ci_df(result_gsi_fixed$CI, "Fixed GSI"),
          ci_df(result_gsi$CI,       "Drawn GSI")) |>
  filter(Stock != "WildSmolts") |>
  ggplot(aes(Estimate, factor(Stock), xmin = LCI, xmax = UCI, colour = Workflow)) +
  geom_pointrange(position = position_dodge(width = 0.5), size = 0.6) +
  scale_colour_manual(values = c("Fixed GSI" = "#2166ac", "Drawn GSI" = "#d73027")) +
  scale_x_continuous(labels = comma) +
  labs(title = "Per-stock smolts: fixed vs drawn GSI (GE fixed) — MY2025",
       x = "Estimated abundance", y = NULL, colour = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", axis.text.y = element_text(hjust = 1))
```

The total is unchanged by GSI draws; per-stock CIs widen (`Width_ratio > 1`). GE
and GSI draws can be supplied together for full propagation. Totals from this
genotyped subset need not match the full-trap results in §6.

---

# Reviewer checklist

| Check | Expected result |
|---|---|
| `gelman.diag()` R-hat (psi, mu_strat, alpha, beta, sigma_psi/strat) | < 1.05 for all |
| GE draws range | All in \[0, 1\], no NaN |
| Stratum-guided weekly GE plot | Weekly GE shrinks toward collapse-stratum line; sparse weeks pulled toward parent; both below NOAA at high spill |
| **Path A vs Path B** | **WildSmolts within ~5% (bootstrap noise) — SCRAPI ≡ SCRAPI2** |
| **Exact-GE control** | **< 1% WildSmolts difference — geDraws pathway neutral** |
| **Degenerate-GSI control** | **< 1% WildSmolts difference — gsiDraws pathway neutral** |
| Path C CI vs Path B | Wider or equal (GE uncertainty propagated) |
| Path C CI asymmetry | Upper tail longer than lower — expected from convexity of 1/GE expansion |
| Fixed vs drawn GSI | Total unchanged; per-stock CIs wider (Width_ratio > 1); LOCLWR/LOSALM largest |
