---
title: "SCRAPI2: Propagating GE and GSI uncertainty in smolt escapement estimation"
subtitle: "Validation against SCOBI::SCRAPI — MY2025 steelhead"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
vignette: >
  %\VignetteIndexEntry{SCRAPI2: GE and GSI uncertainty in smolt escapement estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  eval    = TRUE
)
```

---

# Background

## What SCRAPI does

SCRAPI (Steinhorst & Ackerman) estimates wild smolt escapement past Lower Granite
Dam (LGR) by expanding trap catches to the full passage using a daily sampling
rate. The core formula for daily estimated passage is:

$$\hat{N}_d = \frac{C_d}{R_d \times GE_d}$$

where $C_d$ is the trap count, $R_d$ is the sampling rate (fraction of fish
entering the bypass that are trapped), and $GE_d$ is the guidance efficiency —
the fraction of all downstream migrants that actually enter the juvenile bypass
system. The bootstrap resamples both the passage counts and the trap fish to
propagate uncertainty in $R_d$ and the stock/rear composition into the CIs.

## The GE problem

Historically, $GE_d$ is a **fixed point estimate** taken from a NOAA table
(modelled from spill and fish-behaviour studies). Plugging in a fixed $GE$ means
the bootstrap CIs reflect only uncertainty in trap sampling — not uncertainty in
GE itself. Under post-ROD operations LGR runs at 80–85% spill during peak
migration, where GE is both low and poorly constrained, so a fixed GE can badly
understate the true uncertainty in escapement.

## What SCRAPI2 adds

`SCRAPI2` extends the SCRAPI logic in two ways, used independently or together:

1. **GE draws (`geDraws`)** — instead of a single $GE_d$ per day, supply a
   matrix of posterior draws from a Bayesian hierarchical model
   (`fit_ge_model()` / `generate_ge_draws()`). Each bootstrap iteration uses a
   different column (draw), so GE uncertainty flows through into the CIs. The GE
   model can be fit at any stratum resolution (down to weekly), shrink fine
   strata toward coarser parents, and resolve GE to a daily value that tracks
   daily spill — see Section 4.

2. **GSI draws (`gsiDraws`)** — instead of fixed stock assignments in the trap,
   supply posterior draws from a genetic stock identification model. Each
   bootstrap iteration uses a different draw so genetic uncertainty flows through
   into per-stock CIs. This mirrors the adult-side **EASE** framework: one column
   per posterior sample, one row per fish, keyed by `MasterID`.

**Key guarantee (validated below):** when every column of `geDraws` is identical
(or when `geDraws = NULL` and a fixed `GuidanceEfficiency` column is supplied)
and `gsiDraws = NULL`, `SCRAPI2` reproduces `SCOBI::SCRAPI` to within bootstrap
noise. SCRAPI2 changes *what uncertainty is propagated*, not the underlying
estimation arithmetic.

---

# Install and load

```{r install, eval=FALSE}
remotes::install_github("dsmi313/smoltEASE")
```

```{r libraries}
library(smoltEASE)
library(tidyverse)
library(lubridate)
library(readxl)
library(scales)
library(coda)      # gelman.diag for MCMC convergence
```

```{r config}
B <- 2000   # bootstrap iterations (and GE/GSI draws)

# Date window for the migration year
season_start <- ymd("2025-03-24")
season_end   <- ymd("2025-08-01")
```

---

# Data inputs

All dates are handled as `"%m/%d/%Y"` strings (the trap/FPC convention) and
`SCRAPI2`/`SCRAPI` are called with `dateFormat = "%m/%d/%Y"`.

## Trap, FPC and GSI draws

```{r data-load}
# Pre-formatted trap data: one row per sampled smolt (CollectionDate, Rear,
# GenStock, fwAge, MasterID). Normalise CollectionDate to "%m/%d/%Y".
STHD.trap <- read.csv("MY2025STHD.trapData_formatted.csv")
STHD.trap$CollectionDate <- format(as.Date(STHD.trap$CollectionDate, "%Y-%m-%d"),
                                    "%m/%d/%Y")

# FPC passage data with weekly `Collapse` strata AND the fixed NOAA
# `GuidanceEfficiency` already merged in (one row per passage day).
STHD.FPC <- read.csv("MY2025STHDsmolts_FPCdata_strata_added.csv")
STHD.FPC$SampleEndDate <- format(as.Date(STHD.FPC$SampleEndDate, "%m/%d/%Y"),
                                 "%m/%d/%Y")

# GSI posterior draws: column 1 = MasterID, remaining columns = stock draws.
gsi_draws <- read.delim("OmyLGRJ25C_gsiDraws_final.txt",
                        sep = "\t", header = TRUE, check.names = FALSE)

stopifnot(all(c("Collapse", "GuidanceEfficiency", "SampleEndDate",
                "SampleCount", "SampleRate", "Week") %in% names(STHD.FPC)))
```

## Strata mapping

Week-to-stratum collapse mapping from the `STHD` tab of `Smolt_Strata.xlsx`.
Reused as the passage-bootstrap strata and (Section 4.7) the *parent* grouping
for nested weekly GE. The FPC already carries `Collapse`; this table drives the
GE model and the legacy `Collapse` join for the trap.

```{r strata, R.options=list(width=200)}
strata_raw <- read_xlsx("Smolt_Strata.xlsx", sheet = 1)
sthd_start <- which(strata_raw[[1]] == "WeekNumber" & strata_raw[[2]] == "STHD_n")

sthd_strata <- strata_raw[(sthd_start + 1):nrow(strata_raw), c(1, 3)] |>
  set_names(c("Week", "Collapse")) |>
  filter(!is.na(Week)) |>
  mutate(Week = as.integer(Week), Collapse = as.integer(Collapse))

genstock_by_week(STHD.trap, species = "sthd")   # composition by week (optional QC)
```

## PIT detections for the GE model

GRJ (bypass), GRS (spillway bay 1) and GOJ (Little Goose bypass) first-detections
per tag, restricted to the migration window. This `pit_raw` is the `dat_up`
input to `prep_ge_data()` (tag / site / det_date).

```{r pit-prep}
pit_raw <- read_csv("MY25_STHD.csv") |>
  mutate(
    site     = str_extract(`Site Name`, "^[A-Z]+"),
    obs_date = as_date(parse_date_time(`Obs Time Value`,
                                       orders = c("mdy HM", "mdy HMS")))
  ) |>
  filter(
    `Rear Type Code` %in% c("W", "H", "U"),
    site %in% c("GRJ", "GRS", "GOJ"),
    obs_date >= season_start, obs_date < season_end
  ) |>
  group_by(`Tag Code`, site) |>
  slice_min(obs_date, with_ties = FALSE) |>
  ungroup() |>
  rename(tag = `Tag Code`, det_date = obs_date)
```

## Daily spill (LGR and LGS)

LGR spill is the covariate for $\psi$ (GE) and $p$, and (Section 4.7) drives the
daily GE downscale. LGS spill is the covariate for the GOJ detection probability
$\phi$.

```{r spill}
# LGR daily spill (% of flow)
lgr_spill <- read_csv("DART Spill.csv") |>
  mutate(
    Date      = mdy(date),
    spill.per = as.numeric(spill_kcfs) / as.numeric(outflow_kcfs) * 100,
    spill.per = replace_na(spill.per, 0)
  ) |>
  filter(Date >= season_start, Date < season_end) |>
  select(Date, spill.per)

# LGS daily spill, aggregated from hourly files
lgs_spill <- bind_rows(
  read_csv("spill.all.csv"), read_csv("spill.all2.csv"), read_csv("spill.all3.csv")
) |>
  mutate(dt = mdy_hm(DateTime), Date = as_date(dt),
         spill.per = HourlySpill / HourlyFlow * 100) |>
  filter(Site == "LGS", Date >= season_start, Date < season_end) |>
  group_by(Date) |>
  summarise(spill.per = mean(spill.per, na.rm = TRUE), .groups = "drop")
```

---

# The Bayesian GE model

## What it estimates

The current model is a **multistate mark-recapture model** (after Hance et al.
2020/2024), not the legacy binomial psi-pool. From PIT-tagged fish detected at
GRJ (bypass), GRS (spillway bay 1) and GOJ (Little Goose bypass) it jointly
estimates, per stratum $s$, three probabilities:

- $\psi_s$ — probability a fish passing LGR takes the **juvenile bypass** route.
  **This is guidance efficiency directly:** $GE_s = \psi_s$.
- $p_s$ — probability a spillway fish is **detected at the GRS** bay-1 antenna.
- $\phi_s$ — probability a fish is **detected at GOJ** downstream.

Each detected fish falls into one of five histories, with cell probabilities
(conditioned on being detected somewhere):

$$
\begin{aligned}
\text{(GRJ, no GOJ)}\ &:\ \psi_s (1-\phi_s) \\
\text{(GRJ, GOJ)}\ &:\ \psi_s \phi_s \\
\text{(GRS, no GOJ)}\ &:\ (1-\psi_s)\, p_s (1-\phi_s) \\
\text{(GRS, GOJ)}\ &:\ (1-\psi_s)\, p_s \phi_s \\
\text{(no LGR, GOJ)}\ &:\ (1-\psi_s)(1-p_s)\phi_s
\end{aligned}
$$

The last cell — fish missed by both LGR arrays but recovered at GOJ — is what
makes $p_s$ (and hence $\psi_s$) identifiable. The stratum counts of the five
histories follow a multinomial likelihood. Each parameter is a **hierarchical
logistic regression on spill** ($\psi$ and $p$ on LGR spill, $\phi$ on LGS
spill), with weakly-informative Student-$t$ priors for $\psi/p$ (and $\beta\le0$:
more spill cannot raise guidance) and **informative Normal priors on the $\phi$
coefficients from McCann et al. (2023) Table 8.3**. The hierarchy lets strata be
as fine as weeks; sparse weeks are partially pooled toward the spill regression.

## The JAGS model

This is the single-level model `fit_ge_model()` builds internally (the nested
variant of Section 4.7 replaces the `logit_psi` line with a parent-group mean):

```
model {
  for (s in 1:S) {
    logit_psi[s] ~ dnorm(alpha     + beta     * lgr_spill_std[s], tau_psi)
    logit_p[s]   ~ dnorm(alpha_p   + beta_p   * lgr_spill_std[s], tau_p)
    logit_phi[s] ~ dnorm(alpha_phi + beta_phi * lgs_spill_std[s], tau_phi)
    psi[s] <- ilogit(logit_psi[s])     # psi == GE
    p[s]   <- ilogit(logit_p[s])
    phi[s] <- ilogit(logit_phi[s])

    pi_raw[s,1] <- psi[s]*(1-phi[s])                # GRJ, no GOJ
    pi_raw[s,2] <- psi[s]*phi[s]                     # GRJ, GOJ
    pi_raw[s,3] <- (1-psi[s])*p[s]*(1-phi[s])        # GRS, no GOJ
    pi_raw[s,4] <- (1-psi[s])*p[s]*phi[s]            # GRS, GOJ
    pi_raw[s,5] <- (1-psi[s])*(1-p[s])*phi[s]        # no LGR, GOJ
    pi_obs[s,] <- pi_raw[s,] / sum(pi_raw[s,])       # condition on detection
  }
  for (j in 1:S_lik) n[lik_idx[j],1:5] ~ dmulti(pi_obs[lik_idx[j],1:5], N_seen[lik_idx[j]])

  alpha ~ dt(0, pow(2,-2), 7);  beta ~ dt(0, pow(1,-2), 7) T(,0);  sigma_psi ~ dunif(0.05,3)
  alpha_p ~ dt(0, pow(2,-2), 7); beta_p ~ dt(0, pow(1,-2), 7);     sigma_p   ~ dunif(0.05,3)
  alpha_phi ~ dnorm(-2.60, pow(0.30,-2))           # McCann et al. 2023, Tab 8.3
  beta_phi  ~ dnorm(-1.78, pow(0.25,-2)) T(,0)
  sigma_phi ~ dunif(0.05,1)
}
```

## Prepare the stratum-level GE data

```{r ge-prep}
ge_data <- prep_ge_data(
  dat_up         = pit_raw,
  strat_assign   = sthd_strata,
  spill_data     = lgr_spill,
  lgs_spill_data = lgs_spill,
  species        = "sthd",
  goj_site       = "GOJ"
)

ge_data %>%
  select(stratum, spill_LGR = spill_val, spill_LGS = lgs_spill_val,
         n_pool, n_GRJ = n_GRJ_obs, n_GRS = n_GRS_obs,
         n_h1, n_h2, n_h3, n_h4, n_h5) %>%
  knitr::kable(caption = "Stratum-level inputs to the multistate GE model")
```

## Fit the model and check convergence

```{r ge-fit}
ge_fit <- fit_ge_model(ge_data, max_n = 500, seed = 42)

# R-hat on GE parameters (psi == GE) and the psi regression scalars; want < 1.05
psi_cols <- grep("^psi\\[|^alpha$|^beta$|^sigma_psi$",
                 coda::varnames(ge_fit$samples))
gelman.diag(ge_fit$samples[, psi_cols], autoburnin = FALSE, multivariate = FALSE)
```

## Stratum GE (model posterior) vs NOAA fixed GE

Because $GE_s = \psi_s$, the stratum-level model GE is the $\psi$ posterior. We
compare it to the fixed NOAA GE carried in the FPC, aggregated to the same
strata.

```{r ge-compare-table}
noaa_by_collapse <- STHD.FPC %>%
  group_by(Collapse) %>%
  summarise(NOAA_GE = round(mean(GuidanceEfficiency, na.rm = TRUE), 3),
            .groups = "drop")

psi_mat <- as.matrix(ge_fit$samples)[, grep("^psi\\[",
             coda::varnames(ge_fit$samples)), drop = FALSE]

ge_by_stratum <- data.frame(
  stratum   = ge_fit$obs_strata$stratum,
  spill_LGR = round(ge_fit$obs_strata$spill_val, 1),
  Model_GE  = round(colMeans(psi_mat), 3),
  Model_lo  = round(apply(psi_mat, 2, quantile, 0.025), 3),
  Model_hi  = round(apply(psi_mat, 2, quantile, 0.975), 3)
) %>%
  left_join(noaa_by_collapse, by = c("stratum" = "Collapse"))

knitr::kable(
  ge_by_stratum,
  col.names = c("Stratum", "Spill (%)", "Model GE", "2.5%ile", "97.5%ile",
                "NOAA fixed GE"),
  caption   = "Stratum GE: multistate model posterior (95% CI) vs NOAA fixed"
)
```

```{r ge-compare-plot}
ge_by_stratum %>%
  pivot_longer(c(Model_GE, NOAA_GE), names_to = "source", values_to = "GE") %>%
  ggplot(aes(spill_LGR, GE, colour = source, label = stratum)) +
  geom_line(aes(group = source), linewidth = 0.8) +
  geom_point(size = 3) +
  geom_text(nudge_y = 0.01, show.legend = FALSE, size = 3) +
  scale_colour_manual(values = c(Model_GE = "#d73027", NOAA_GE = "#4dac26"),
                      labels = c(Model_GE = "Model GE (psi)",
                                 NOAA_GE  = "NOAA fixed GE")) +
  labs(title = "GE by stratum: multistate model vs NOAA",
       x = "Mean LGR spill (%)", y = "Guidance efficiency", colour = NULL) +
  theme_minimal(base_size = 13)
```

## Generate GE draws for SCRAPI2

Each `boot_*` column is one posterior draw of daily GE; `SCRAPI2` consumes one
column per bootstrap iteration (the same mechanism as GSI draws).

```{r ge-draws}
ge_draws <- generate_ge_draws(
  ge_fit       = ge_fit,
  strat_assign = sthd_strata,
  pass_dates   = as.Date(STHD.FPC$SampleEndDate, "%m/%d/%Y"),
  B            = B
)

stopifnot(nrow(ge_draws) == nrow(STHD.FPC))
stopifnot(all(unlist(ge_draws[, -1]) >= 0), all(unlist(ge_draws[, -1]) <= 1))
cat("Posterior mean GE by day:\n"); summary(rowMeans(ge_draws[, -1]))
```

## Finer resolution: weekly strata, nested shrinkage, daily spill

The same functions support a finer, more defensible GE without changing
`SCRAPI2`. All are opt-in.

- **Weekly strata** — pass a `strat_assign` with `Collapse = Week`.
- **Nested shrinkage** — pass your `Week`/`Collapse` table as `parent_strata`;
  each week's $\psi$ shrinks toward its collapse-stratum mean, so sparse weeks
  borrow strength instead of producing high-leverage near-zero GE. Data-driven,
  no floor.
- **Daily spill** — pass `daily_spill` to `generate_ge_draws()`; GE follows that
  day's spill via the fitted regression, anchored by each draw's stratum
  deviation (days share $\alpha,\beta,\varepsilon$ within a draw, so no spurious
  independent daily noise).

```{r ge-weekly-nested-daily, eval=FALSE}
sthd_weekly <- transform(sthd_strata, Collapse = Week)        # one stratum per week

ge_data_wk <- prep_ge_data(
  dat_up = pit_raw, strat_assign = sthd_weekly, spill_data = lgr_spill,
  lgs_spill_data = lgs_spill, species = "sthd", goj_site = "GOJ",
  parent_strata = sthd_strata)                                # weeks nested in strata
ge_fit_wk <- fit_ge_model(ge_data_wk, max_n = 500, seed = 42) # nested model auto-detected

ge_draws_daily <- generate_ge_draws(
  ge_fit_wk, strat_assign = sthd_weekly,
  pass_dates = as.Date(STHD.FPC$SampleEndDate, "%m/%d/%Y"),
  B = B, daily_spill = lgr_spill)                             # daily-resolved GE
# ge_draws_daily plugs into SCRAPI2 exactly like ge_draws.
```

---

# Three-way comparison

Three workflows on the same MY2025 steelhead data; same fish, same passage.

| Workflow | GE source | Tool |
|---|---|---|
| **A — Legacy** | NOAA `GuidanceEfficiency` (in FPC) | `SCOBI::SCRAPI` |
| **B — SCRAPI2 fixed GE** | Same NOAA column, no uncertainty | `SCRAPI2` (`geDraws = NULL`) |
| **C — SCRAPI2 model GE** | Bayesian posterior draws | `SCRAPI2` (`geDraws = ge_draws`) |

Paths A and B should produce near-identical point estimates and CI widths
(validating that `SCRAPI2` did not change the SCRAPI logic). Path C should have
**wider** CIs because GE uncertainty is now propagated.

## Helper: add Collapse to the trap for legacy SCOBI::SCRAPI

`SCOBI::SCRAPI` needs `Collapse` in both data frames. The FPC already has it; the
trap does not, so map it on from `sthd_strata`.

```{r add-collapse-helper}
add_collapse <- function(df, date_col, strata_tbl) {
  d              <- as.Date(df[[date_col]], "%m/%d/%Y")
  df$Week        <- isoweek(d)
  df             <- left_join(df, strata_tbl, by = "Week")
  df[[date_col]] <- format(d, "%m/%d/%Y")
  df
}
```

## Path A — Legacy `SCOBI::SCRAPI`

```{r path-a-legacy}
STHD.trap_legacy <- add_collapse(STHD.trap, "CollectionDate", sthd_strata)
stopifnot(!anyNA(STHD.trap_legacy$Collapse), !anyNA(STHD.FPC$Collapse))

SCOBI::SCRAPI(
  smoltData   = STHD.trap_legacy,
  passageData = STHD.FPC,                 # carries Collapse + GuidanceEfficiency
  Primary     = "GenStock",
  dateFormat  = "%m/%d/%Y",
  Run         = "legacy_",
  B           = B
)

# SCOBI::SCRAPI returns NULL and writes CSV; read it back, dropping footer rows.
legacy_ci <- read.csv("legacy_CIs.csv", skip = 1,
                      header = TRUE, row.names = 1, check.names = FALSE) %>%
  filter(!grepl("SampleSize", rownames(.)))
```

## Path B — `SCRAPI2` with fixed GE (validation run)

```{r path-b-fixed-ge}
result_fixed <- SCRAPI2(
  smoltData   = STHD.trap,
  passageData = STHD.FPC,                 # fixed NOAA GuidanceEfficiency, geDraws = NULL
  Primary     = "GenStock",
  strat       = "Week", dat = "SampleEndDate", tally = "SampleCount",
  samrate     = "SampleRate", guidance = "GuidanceEfficiency", collaps = "Collapse",
  RTYPE       = "W", alph = 0.05, dateFormat = "%m/%d/%Y",
  Run         = "fixed_", B = B
)
```

## Path C — `SCRAPI2` with Bayesian GE draws

```{r path-c-model-ge}
result_model <- SCRAPI2(
  smoltData   = STHD.trap,
  passageData = STHD.FPC,
  geDraws     = ge_draws,
  Primary     = "GenStock",
  strat       = "Week", dat = "SampleEndDate", tally = "SampleCount",
  samrate     = "SampleRate", guidance = "GuidanceEfficiency", collaps = "Collapse",
  RTYPE       = "W", alph = 0.05, dateFormat = "%m/%d/%Y",
  Run         = "model_", B = B
)
```

---

# Results

## Wild smolt totals

```{r results-table}
ci_df <- function(mat, label) {
  as.data.frame(mat) %>% rownames_to_column("Stock") %>% mutate(Workflow = label)
}

tbl_legacy <- legacy_ci %>% rownames_to_column("Stock") %>%
  mutate(Workflow = "A: SCOBI legacy")
tbl_fixed  <- ci_df(result_fixed$CI, "B: SCRAPI2 fixed GE")
tbl_model  <- ci_df(result_model$CI, "C: SCRAPI2 model GE")

results_all <- bind_rows(tbl_legacy, tbl_fixed, tbl_model)

results_all %>%
  filter(Stock == "WildSmolts") %>%
  mutate(`CI half-width (%)` = round((UCI - LCI) / (2 * Estimate) * 100, 1)) %>%
  select(Workflow, Estimate, LCI, UCI, `CI half-width (%)`) %>%
  knitr::kable(caption = "Wild steelhead smolt estimates — three workflows",
               format.args = list(big.mark = ","))
```

## Forest plot — all stocks

```{r forest-plot, fig.width=10, fig.height=5}
results_all %>%
  filter(Stock != "WildSmolts") %>%
  mutate(Stock = factor(Stock)) %>%
  ggplot(aes(x = Estimate, xmin = LCI, xmax = UCI, y = Stock, colour = Workflow)) +
  geom_pointrange(position = position_dodge(width = 0.5), size = 0.6) +
  scale_colour_manual(values = c("A: SCOBI legacy" = "#555555",
                                 "B: SCRAPI2 fixed GE" = "#2166ac",
                                 "C: SCRAPI2 model GE" = "#d73027")) +
  scale_x_continuous(labels = scales::comma) +
  labs(title = "Wild steelhead smolts by genetic stock — MY2025",
       x = "Estimated abundance", y = NULL, colour = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", axis.text.y = element_text(hjust = 1))
```

## CI width comparison — WildSmolts

```{r ci-width-plot, fig.width=6, fig.height=4}
results_all %>%
  filter(Stock == "WildSmolts") %>%
  mutate(half_width = (UCI - LCI) / 2, Workflow = fct_inorder(Workflow)) %>%
  ggplot(aes(x = Workflow, y = half_width, fill = Workflow)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  scale_fill_manual(values = c("A: SCOBI legacy" = "#555555",
                               "B: SCRAPI2 fixed GE" = "#2166ac",
                               "C: SCRAPI2 model GE" = "#d73027")) +
  labs(title = "CI half-width — WildSmolts", x = NULL, y = "Half-width (fish)") +
  theme_minimal(base_size = 13)
```

The model-GE CI (Path C) should be wider than the fixed-GE CI (Paths A/B) — the
extra width is the uncertainty from not knowing GE exactly.

---

# GSI uncertainty: fixed vs drawn

GE uncertainty widens the **total** wild-smolt CI. GSI uncertainty instead widens
the **per-stock** CIs: it does not change how many wild smolts there are (rear
composition and passage are untouched), only how confidently we split them among
genetic stocks. To isolate the genetic effect we hold GE **fixed**, restrict to
the genotyped fish, and compare:

| Workflow | Stock assignment | Tool |
|---|---|---|
| **Fixed GSI** | observed `GenStock`, treated as known | `SCRAPI2` (`gsiDraws = NULL`) |
| **Drawn GSI** | GSI posterior draws, one per bootstrap | `SCRAPI2` (`gsiDraws = gsi_draws`) |

## Restrict the trap to genotyped fish

`SCRAPI2` requires every `smoltData` fish to appear in the draws, so we keep only
trap fish with a posterior (fish without a valid BiosamplesID were not
genotyped).

```{r gsi-load}
gsi_ids       <- gsi_draws[[1]]
STHD.trap.gsi <- STHD.trap[STHD.trap$MasterID %in% gsi_ids, ]
cat(sprintf("Trap fish: %d total, %d with GSI draws, %d excluded\n",
            nrow(STHD.trap), nrow(STHD.trap.gsi),
            nrow(STHD.trap) - nrow(STHD.trap.gsi)))
dim(gsi_draws)           # n_fish x (n_draws + 1)
```

## Fixed vs drawn GSI (same fish, same fixed GE)

```{r gsi-runs}
# Fixed GSI: observed GenStock as known
result_gsi_fixed <- SCRAPI2(
  smoltData = STHD.trap.gsi, passageData = STHD.FPC, Primary = "GenStock",
  strat = "Week", dat = "SampleEndDate", tally = "SampleCount",
  samrate = "SampleRate", guidance = "GuidanceEfficiency", collaps = "Collapse",
  RTYPE = "W", alph = 0.05, dateFormat = "%m/%d/%Y", Run = "gsifixed_", B = B)

# Drawn GSI: posterior draws, one per bootstrap iteration
result_gsi <- SCRAPI2(
  smoltData = STHD.trap.gsi, passageData = STHD.FPC, Primary = "GenStock",
  gsiDraws = gsi_draws, fishID = "MasterID",
  strat = "Week", dat = "SampleEndDate", tally = "SampleCount",
  samrate = "SampleRate", guidance = "GuidanceEfficiency", collaps = "Collapse",
  RTYPE = "W", alph = 0.05, dateFormat = "%m/%d/%Y", Run = "gsidrawn_", B = B)
```

## The total is unchanged; per-stock CIs widen

GSI draws only reassign each fish's stock — not its rear type or the passage
counts — so the wild **total** (and its CI) is identical between the two runs.

```{r gsi-total-check}
cat(sprintf("WildSmolts total  fixed GSI: %s   drawn GSI: %s\n",
            format(result_gsi_fixed$CI["WildSmolts", "Estimate"], big.mark = ","),
            format(result_gsi$CI["WildSmolts",       "Estimate"], big.mark = ",")))
```

```{r gsi-compare-table}
gsi_tab <- bind_rows(ci_df(result_gsi_fixed$CI, "Fixed"),
                     ci_df(result_gsi$CI,       "Drawn")) %>%
  filter(Stock != "WildSmolts") %>%
  transmute(Stock, Workflow, Estimate = round(Estimate),
            HalfWidth = round((UCI - LCI) / 2)) %>%
  pivot_wider(names_from = Workflow, values_from = c(Estimate, HalfWidth)) %>%
  mutate(Width_ratio = round(HalfWidth_Drawn / HalfWidth_Fixed, 2))

knitr::kable(
  gsi_tab,
  caption = paste("Per-stock smolts: fixed vs drawn GSI (genotyped fish, GE",
                  "held fixed). Width_ratio > 1 = genetic uncertainty widens the CI."),
  format.args = list(big.mark = ","))
```

```{r gsi-forest, fig.width=10, fig.height=5}
bind_rows(ci_df(result_gsi_fixed$CI, "Fixed GSI"),
          ci_df(result_gsi$CI,       "Drawn GSI")) %>%
  filter(Stock != "WildSmolts") %>%
  ggplot(aes(x = Estimate, y = Stock, xmin = LCI, xmax = UCI, colour = Workflow)) +
  geom_pointrange(position = position_dodge(width = 0.5), size = 0.6) +
  scale_colour_manual(values = c("Fixed GSI" = "#2166ac", "Drawn GSI" = "#d73027")) +
  scale_x_continuous(labels = scales::comma) +
  labs(title = "Per-stock smolts: fixed vs drawn GSI (GE fixed) — MY2025",
       x = "Estimated abundance", y = NULL, colour = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", axis.text.y = element_text(hjust = 1))
```

The wild **total** matches between the two runs while the **per-stock** CIs are
wider under drawn GSI — `Width_ratio > 1` quantifies the genetic uncertainty
added per stock. GE and GSI draws can be supplied together for full propagation
(`geDraws = ...` **and** `gsiDraws = ...` in one call). These per-stock totals are
on the genotyped subset, so they need not match the full-trap GE comparison.

---

# Control checks (pathway neutrality)

Both the `geDraws` and `gsiDraws` pathways must be **numerically neutral** when
the draws carry no uncertainty — the strict identity test that SCRAPI2 did not
alter the SCRAPI arithmetic.

## Exact-GE draws

Every draw equals the fixed NOAA GE; with `gsiDraws = NULL` the estimate must
match the fixed-GE run to within bootstrap noise.

```{r control-exact-ge, results='hold'}
ge0 <- pmin(pmax(as.numeric(STHD.FPC$GuidanceEfficiency), 1e-6), 1 - 1e-6)
ge_draws_exact <- data.frame(
  SampleEndDate = STHD.FPC$SampleEndDate,
  matrix(rep(ge0, B), nrow = length(ge0), ncol = B), check.names = FALSE)

result_exact <- SCRAPI2(
  smoltData = STHD.trap, passageData = STHD.FPC, geDraws = ge_draws_exact,
  Primary = "GenStock", strat = "Week", dat = "SampleEndDate", tally = "SampleCount",
  samrate = "SampleRate", guidance = "GuidanceEfficiency", collaps = "Collapse",
  RTYPE = "W", alph = 0.05, dateFormat = "%m/%d/%Y", Run = "exact_", B = B)

est_fixed <- result_fixed$CI["WildSmolts", "Estimate"]
est_exact <- result_exact$CI["WildSmolts", "Estimate"]
cat(sprintf("Fixed GE: %s   exact-draw: %s   diff: %.2f%%\n",
            format(est_fixed, big.mark = ","), format(est_exact, big.mark = ","),
            abs(est_fixed - est_exact) / est_fixed * 100))
stopifnot(abs(est_fixed - est_exact) / est_fixed < 0.01)
```

## Degenerate GSI draws

`sim_gsi_draws(method = "fixed")` repeats the **observed** `GenStock` in every
column — real assignments, no genetic uncertainty — so it must reproduce the
`gsiDraws = NULL` run (`result_gsi_fixed`). Same genotyped subset.

```{r control-fixed-gsi, results='hold'}
gsi_degen <- sim_gsi_draws(STHD.trap.gsi, Primary = "GenStock",
                           fishID = "MasterID", B = B, method = "fixed")

result_gsi_degen <- SCRAPI2(
  smoltData = STHD.trap.gsi, passageData = STHD.FPC, gsiDraws = gsi_degen,
  fishID = "MasterID", Primary = "GenStock",
  strat = "Week", dat = "SampleEndDate", tally = "SampleCount",
  samrate = "SampleRate", guidance = "GuidanceEfficiency", collaps = "Collapse",
  RTYPE = "W", alph = 0.05, dateFormat = "%m/%d/%Y", Run = "gsidegen_", B = B)

gsi_ctrl <- merge(
  data.frame(Stock = rownames(result_gsi_fixed$CI), null_est = result_gsi_fixed$CI[, "Estimate"]),
  data.frame(Stock = rownames(result_gsi_degen$CI), degen_est = result_gsi_degen$CI[, "Estimate"]),
  by = "Stock")
gsi_ctrl$pct_diff <- round(abs(gsi_ctrl$null_est - gsi_ctrl$degen_est) /
                             pmax(gsi_ctrl$null_est, 1) * 100, 2)
knitr::kable(gsi_ctrl,
             caption = "GSI pathway control: gsiDraws = NULL vs degenerate (observed-stock) draws",
             format.args = list(big.mark = ","))

w_null  <- result_gsi_fixed$CI["WildSmolts", "Estimate"]
w_degen <- result_gsi_degen$CI["WildSmolts", "Estimate"]
cat(sprintf("WildSmolts  null: %s   degenerate-draw: %s   diff: %.2f%%\n",
            format(w_null, big.mark = ","), format(w_degen, big.mark = ","),
            abs(w_null - w_degen) / w_null * 100))
stopifnot(abs(w_null - w_degen) / w_null < 0.01)
```

---

# Reviewer checklist

| Check | Expected result |
|---|---|
| `gelman.diag()` R-hat (psi, alpha, beta, sigma_psi; + p/phi, mu_strat if nested) | < 1.05 for all monitored parameters |
| GE draws range | All values in \[0, 1\] |
| Stratum GE vs NOAA | Model GE generally below NOAA at high spill (post-ROD) |
| **Path A vs Path B point estimates** | **Within ~5% (bootstrap noise, same seed) — SCRAPI ≡ SCRAPI2** |
| **Exact-GE control check** | **< 1% difference in WildSmolts estimate — geDraws pathway is neutral** |
| **Degenerate-GSI control** | **< 1% difference in WildSmolts estimate — gsiDraws pathway is neutral** |
| Path C CI half-width vs Path B | Wider or equal — never narrower (GE uncertainty) |
| Fixed vs drawn GSI: WildSmolts total | Matches within bootstrap noise (GSI doesn't change the total) |
| Fixed vs drawn GSI: per-stock CIs | Wider under drawn GSI (`Width_ratio` > 1) |
| Weekly + nested fit (if used) | Data-rich weeks unchanged; sparse weeks shrink toward stratum mean |

---
